---
title: "Simulation design and initialization (supplementary parameter sweep)"
format: html
execute:
  echo: true
  message: true
  warning: true
---

This document generates dispersal matrices, planting layouts, and NEMO-age INI files for a **supplementary parameter sweep** across carrying capacity (**K**) and competition (**B**).\
Workflow:

1.  Write **burn-in** INIs for every K×B combination (and a run script).
2.  After burn-in has been executed, parse each burn-in `.txt` to extract the **baseline** per-patch stage structure at generation {"500"}.
3.  For each planting scenario layout, create **scenario-specific stage init files** by clearing + planting Orientalis patches on top of the corresponding baseline.
4.  Write **run** INIs for all scenarios × K×B combinations (and a run script).

> Everything is written under `supplementary/`.

## Packages

```{r}
#| label: packages
#| echo: true

suppressPackageStartupMessages({
  library(terra)
  library(data.table)
  library(ggplot2)
  library(cowplot)
})
```

## Control

```{r}
#| label: control
#| echo: true

# ---- Template INI (single clean template used for burn-in + run)
TEMPLATE_INI <- "ini_files/templates/nemoage_template.ini"
stopifnot(file.exists(TEMPLATE_INI))

# ---- All outputs live in ./supplementary/
SUPP_ROOT <- "supplementary"

OUT <- list(
  input   = file.path(SUPP_ROOT, "input_files"),
  ini     = file.path(SUPP_ROOT, "ini_files"),
  results = file.path(SUPP_ROOT, "results"),
  scripts = file.path(SUPP_ROOT, "scripts")
)

DIRS <- list(
  grid       = file.path(OUT$input, "grid"),
  cfg        = file.path(OUT$input, "configs_txt"),
  quanti     = file.path(OUT$input, "quanti_init_freq"),
  stage      = file.path(OUT$input, "patch_init_stage_size"),
  disperse   = file.path(OUT$input, "disperse"),
  plots      = file.path(OUT$input, "plots"),
  ini_burnin = file.path(OUT$ini, "burnin"),
  ini_run    = file.path(OUT$ini, "run")
)

for (d in c(OUT, DIRS)) dir.create(d, showWarnings = FALSE, recursive = TRUE)

# ---- Landscape
GRID_RES <- 4      # meters (patch = 4x4 m)
EXTENT_M <- 100    # 100x100 m
N        <- EXTENT_M / GRID_RES  # 25
CRS      <- "EPSG:3035"
GRID_SHP <- file.path(DIRS$grid, "Grid_4x4m_100x100m.shp")

# ---- Scenario design
PROP_LEVELS    <- 0.25
CONFIG_LEVELS  <- "dispersed"
N_LAYOUT_RUNS  <- 1  # "biological replicates" per (config × proportion)

# ---- Planting scheme for Orientalis
PLANTING_SCHEME <- c(0, 20, 0, 0)  # stage0..stage3

# ---- Burn-in + run settings
BURNIN_GENS <- 500
RUN_GENS    <- 500

K_VALUES <- c(20, 30, 40)
B_VALUES <- c(0.01, 0.02, 0.03, 0.04, 0.05, 0.06)

# Write stats + quanti at explicit times (include gen 2)
LOG_EVERY <- 10
LOG_TIMES <- c(2, seq(LOG_EVERY, RUN_GENS, by = LOG_EVERY))

# NEMO replicates
NEMO_REPS_BURNIN <- 1
NEMO_REPS_RUN    <- 20

# ---- Dispersal (fixed)
# SEED
SEED_D_MEAN <- 5
SEED_B      <- 1.3
SEED_D_T    <- 50

# POLLEN
POLLEN_D_MEAN <- 30
POLLEN_B      <- 1
POLLEN_D_T    <- 100

# ---- Layout controls
DISPERSED_STRENGTH <- 1.20
CLUSTER_JITTER     <- 0.15
MULTI_N_CLUSTERS   <- 4
TRANSECTS_MIN      <- 2
TRANSECTS_MAX      <- 6

# ---- Base seed (deterministic layout generation)
SEED_BASE <- 4022026

# helper for B tag in file/dir names
fmt_b <- function(b) {
  s <- format(b, scientific = FALSE, trim = TRUE)
  s <- sub("0+$", "", s)
  s <- sub("\\.$", "", s)
  s
}
```

## Grid

```{r}
#| label: grid
#| echo: true

make_or_load_grid_strict <- function(shp_path, extent_m = 100, res = 4, crs = "EPSG:3035") {
  if (file.exists(shp_path)) return(vect(shp_path))

  r <- rast(extent = ext(0, extent_m, 0, extent_m), resolution = res, crs = crs)
  r[] <- 1:ncell(r)  # row-wise patchID in raster order

  grid_poly <- as.polygons(r, dissolve = FALSE)
  names(grid_poly)[names(grid_poly) == "lyr.1"] <- "patchID"
  writeVector(grid_poly, shp_path, filetype = "ESRI Shapefile", overwrite = TRUE)

  grid_poly
}

grid <- make_or_load_grid_strict(GRID_SHP, EXTENT_M, GRID_RES, CRS)
stopifnot(nrow(grid) == N * N)

cent <- centroids(grid)

coords <- as.data.frame(crds(cent))
names(coords) <- c("x", "y")
coords$patchID <- grid$patchID

coords$row <- ((coords$patchID - 1) %/% N) + 1
coords$col <- ((coords$patchID - 1) %%  N) + 1
coords$y_plot <- N - coords$row + 1
coords$x_plot <- coords$col

patch_ids <- coords$patchID
n_patches <- length(patch_ids)

dist_mat <- as.matrix(distance(cent))  # meters
```

## NEMO writers

```{r}
#| label: nemo-writers
#| echo: true

write.matrix.nemo <- function(mat, outfile) {
  cat("{", file = outfile)
  for (i in seq_len(nrow(mat))) {
    cat("{", file = outfile, append = TRUE)
    cat(mat[i, ], sep = ",", file = outfile, append = TRUE)
    cat("}\n", file = outfile, append = TRUE)
  }
  cat("}\n", file = outfile, append = TRUE)
}

# Read NEMO matrix format back in (needed for baseline stage files)
read.matrix.nemo <- function(path, nrow_expected = NULL, ncol_expected = NULL) {
  x <- readLines(path, warn = FALSE)
  x <- x[grepl("^\\s*\\{", x) | grepl("^\\s*\\{.*\\}\\s*$", x)]
  x <- x[grepl("\\{.*\\}", x)]
  rows <- lapply(x, function(line) {
    nums <- gsub("[\\{\\}\\s]", "", line)
    nums <- nums[nzchar(nums)]
    as.numeric(strsplit(nums, ",")[[1]])
  })
  mat <- do.call(rbind, rows)
  if (!is.null(nrow_expected) && nrow(mat) != nrow_expected) {
    stop("Row mismatch in ", path, ": got ", nrow(mat), " expected ", nrow_expected)
  }
  if (!is.null(ncol_expected) && ncol(mat) != ncol_expected) {
    stop("Col mismatch in ", path, ": got ", ncol(mat), " expected ", ncol_expected)
  }
  mat
}
```

## Dispersal matrices (fixed)

```{r}
#| label: dispersal
#| echo: true

exp_power_kernel <- function(d, b, x) {
  a <- d * exp(lgamma(2 / b) - lgamma(3 / b))
  b * exp(-(x^b / a^b)) / (2 * pi * a^2 * exp(lgamma(2 / b)))
}

build.reduced.dispersal.matrices <- function(num_patch, distance_matrix, d, b, d_thresh) {
  stopifnot(nrow(distance_matrix) == num_patch, ncol(distance_matrix) == num_patch)

  rate_full <- matrix(0, nrow = num_patch, ncol = num_patch)
  conn      <- matrix(NA_integer_, nrow = num_patch, ncol = num_patch)
  rate_red  <- matrix(NA_real_,    nrow = num_patch, ncol = num_patch)

  for (i in 1:num_patch) {
    rate_full[i, ] <- exp_power_kernel(d, b, distance_matrix[i, ])
    ord <- order(rate_full[i, ], decreasing = TRUE)

    conn[i, ]     <- ord
    rate_red[i, ] <- rate_full[i, ord]

    to_remove <- which(rate_red[i, ] < d_thresh)
    if (length(to_remove)) {
      rate_red[i, to_remove] <- NA
      conn[i, to_remove]     <- NA
    }

    s <- sum(rate_red[i, ], na.rm = TRUE)
    if (s > 0) rate_red[i, !is.na(rate_red[i, ])] <- rate_red[i, !is.na(rate_red[i, ])] / s
  }

  list(dispersal_matrix = rate_full, connectivity_matrix = conn, rate_matrix = rate_red)
}

write.red.dispersal.matrix <- function(connectivity_matrix, rate_matrix, conn_file, rate_file) {
  # connectivity
  cat("{", file = conn_file)
  for (i in 1:nrow(connectivity_matrix)) {
    cat("{", file = conn_file, append = TRUE)
    cat(i, file = conn_file, append = TRUE)
    if (!all(is.na(connectivity_matrix[i, ]))) {
      non_na <- connectivity_matrix[i, which(!is.na(connectivity_matrix[i, ]))]
      non_na <- non_na[non_na != i]
      if (length(non_na)) {
        cat(",", file = conn_file, append = TRUE)
        cat(non_na, sep = ",", file = conn_file, append = TRUE)
      }
    }
    cat("}\n", file = conn_file, append = TRUE)
  }
  cat("}\n", file = conn_file, append = TRUE)

  # rate
  cat("{", file = rate_file)
  for (i in 1:nrow(rate_matrix)) {
    cat("{", file = rate_file, append = TRUE)
    if (all(is.na(rate_matrix[i, ]))) {
      cat(1, file = rate_file, append = TRUE)
    } else {
      cat(rate_matrix[i, which(!is.na(rate_matrix[i, ]))], sep = ",", file = rate_file, append = TRUE)
    }
    cat("}\n", file = rate_file, append = TRUE)
  }
  cat("}\n", file = rate_file, append = TRUE)
}

seed_thresh   <- exp_power_kernel(SEED_D_MEAN,   SEED_B,   SEED_D_T)
pollen_thresh <- exp_power_kernel(POLLEN_D_MEAN, POLLEN_B, POLLEN_D_T)

seed_mats <- build.reduced.dispersal.matrices(n_patches, dist_mat, SEED_D_MEAN, SEED_B, seed_thresh)
pol_mats  <- build.reduced.dispersal.matrices(n_patches, dist_mat, POLLEN_D_MEAN, POLLEN_B, pollen_thresh)

seed_conn_file   <- file.path(DIRS$disperse, sprintf("Seed_connectivity_matrix_d%s.txt", SEED_D_MEAN))
seed_rate_file   <- file.path(DIRS$disperse, sprintf("Seed_rate_matrix_d%s.txt", SEED_D_MEAN))
pollen_conn_file <- file.path(DIRS$disperse, sprintf("Pollen_connectivity_matrix_d%s.txt", POLLEN_D_MEAN))
pollen_rate_file <- file.path(DIRS$disperse, sprintf("Pollen_rate_matrix_d%s.txt", POLLEN_D_MEAN))

write.red.dispersal.matrix(seed_mats$connectivity_matrix, seed_mats$rate_matrix, seed_conn_file, seed_rate_file)
write.red.dispersal.matrix(pol_mats$connectivity_matrix,  pol_mats$rate_matrix,  pollen_conn_file, pollen_rate_file)

message("Wrote dispersal matrices to: ", DIRS$disperse)
```

## Burn-in initialization inputs (shared by all K×B)

```{r}
#| label: burnin-inputs
#| echo: true

# sylvatica-only quanti init: all zeros
q_burnin <- matrix(0, nrow = n_patches, ncol = 1)
q_burnin_file <- file.path(DIRS$quanti, "quanti_init_file_burnin_sylv.txt")
write.matrix.nemo(q_burnin, q_burnin_file)

# burn-in stage init: some adults + uniform seedlings
build_patch_init_stage_size_burnin <- function(n_patches, adult_occupancy = 0.20, stage1_seedlings = 5, seed = 1) {
  set.seed(seed)
  mat <- matrix(0, nrow = n_patches, ncol = 4)
  colnames(mat) <- c("stage0","stage1","stage2","stage3")

  adults <- rep(0, n_patches)
  idx <- sample(seq_len(n_patches), size = as.integer(round(adult_occupancy * n_patches)))
  adults[idx] <- 1

  mat[, "stage3"] <- adults
  mat[, "stage1"] <- stage1_seedlings
  mat
}

burnin_stage <- build_patch_init_stage_size_burnin(
  n_patches,
  adult_occupancy = 0.20,
  stage1_seedlings = 5,
  seed = SEED_BASE
)
burnin_stage_file <- file.path(DIRS$stage, "patch_init_stage_size_burnin_sylv.txt")
write.matrix.nemo(burnin_stage, burnin_stage_file)

message("Burn-in init files written:\n- ", q_burnin_file, "\n- ", burnin_stage_file)
```

## INI templating helpers

```{r}
#| label: ini-helpers
#| echo: true

read_ini_lines <- function(path) readLines(path, warn = FALSE)

set_ini_param <- function(lines, key, value) {
  rx <- paste0("^\\s*", key, "\\b")
  idx <- grep(rx, lines)
  newline <- paste(key, value)

  if (length(idx) == 0) return(c(lines, newline))

  lines[idx[1]] <- newline
  if (length(idx) > 1) lines <- lines[-idx[-1]]
  lines
}
```

## Write burn-in INIs for all K×B

```{r}
#| label: write-burnin-inis
#| echo: true

burn_template <- read_ini_lines(TEMPLATE_INI)

# fixed parts
burn_template <- set_ini_param(burn_template, "run_mode", "overwrite")
burn_template <- set_ini_param(burn_template, "replicates", NEMO_REPS_BURNIN)
burn_template <- set_ini_param(burn_template, "generations", BURNIN_GENS)

burn_template <- set_ini_param(burn_template, "quanti_init_freq", paste0("&", q_burnin_file))
burn_template <- set_ini_param(burn_template, "patch_init_stage_size", paste0("&", burnin_stage_file))

burn_template <- set_ini_param(burn_template, "seed_disperse_connectivity_matrix", paste0("&", seed_conn_file))
burn_template <- set_ini_param(burn_template, "seed_disperse_reduced_matrix",       paste0("&", seed_rate_file))
burn_template <- set_ini_param(burn_template, "breed_disperse_connectivity_matrix_mal", paste0("&", pollen_conn_file))
burn_template <- set_ini_param(burn_template, "breed_disperse_reduced_matrix_mal",       paste0("&", pollen_rate_file))

# logging
burn_template <- set_ini_param(burn_template, "stat_log_time",
                              paste0("{{", paste(LOG_TIMES, collapse = ","), "}}"))
burn_template <- set_ini_param(burn_template, "quanti_dir", "quanti")
burn_template <- set_ini_param(burn_template, "quanti_logtime",
                              paste0("{{", paste(LOG_TIMES, collapse = ","), "}}"))

# ensure txt/bygen written
burn_template <- set_ini_param(burn_template, "save_files", 6)

burnin_manifest <- data.table()

for (K in K_VALUES) {
  for (B in B_VALUES) {
    btag <- fmt_b(B)
    burnin_name <- sprintf("burnin_k%d_b%s", K, btag)
    burnin_root <- file.path(OUT$results, "burnin", sprintf("k%d_b%s", K, btag))
    dir.create(burnin_root, showWarnings = FALSE, recursive = TRUE)

    ini <- burn_template

    # deterministic-but-unique seed per (K,B)
    seed_combo <- SEED_BASE + K * 1000 + as.integer(round(B * 1e6))
    ini <- set_ini_param(ini, "random_seed", seed_combo)

    ini <- set_ini_param(ini, "root_dir", burnin_root)
    ini <- set_ini_param(ini, "filename", burnin_name)
    ini <- set_ini_param(ini, "logfile",  paste0(burnin_name, "_log"))

    ini <- set_ini_param(ini, "patch_nbfem", K)
    ini <- set_ini_param(ini, "regulation_by_competition", B)

    burnin_ini <- file.path(DIRS$ini_burnin, paste0(burnin_name, ".ini"))
    writeLines(ini, burnin_ini)

    burnin_txt <- file.path(burnin_root, paste0(burnin_name, ".txt"))

    burnin_manifest <- rbind(
      burnin_manifest,
      data.table(K = K, B = B, btag = btag,
                 burnin_ini = burnin_ini,
                 burnin_root = burnin_root,
                 burnin_txt = burnin_txt,
                 seed = seed_combo)
    )
  }
}

burnin_manifest_file <- file.path(OUT$scripts, "burnin_manifest.tsv")
fwrite(burnin_manifest, burnin_manifest_file, sep = "\t")
message("Wrote burn-in INIs: ", nrow(burnin_manifest))
message("Burn-in manifest: ", burnin_manifest_file)
```

## Script to run all burn-ins

```{r}
#| label: write-burnin-run-script
#| echo: true

cmds <- sprintf("./nemoage %s", burnin_manifest$burnin_ini)

sh_burn <- file.path(OUT$scripts, "run_all_burnin.sh")
writeLines(c("#!/usr/bin/env bash", "set -euo pipefail", "", cmds), sh_burn)
Sys.chmod(sh_burn, mode = "0755")

message("Wrote: ", sh_burn)
```

Run burn-ins:

``` bash
./supplementary/scripts/run_all_burnin.sh
```

## Extract baseline stage matrices from each burn-in output

```{r}
#| label: baseline-from-burnins
#| echo: true

extract_baseline_stage_matrix <- function(txt_path, n_patches, gen = 500, rep = 1) {
  stopifnot(file.exists(txt_path))
  hdr <- names(fread(txt_path, nrows = 0))

  # some runs use off.* for stage0
  hdr2 <- gsub("^off\\.", "a0.", hdr)

  stage_cols <- lapply(0:3, function(s) {
    grep(paste0("^a", s, "\\.fem\\.p\\d+$"), hdr2, value = TRUE)
  })
  if (any(lengths(stage_cols) == 0)) {
    stop("Could not find per-patch female columns a0..a3 in: ", txt_path)
  }

  cols_needed <- c("replicate", "generation", unlist(stage_cols))
  dt <- fread(txt_path, select = cols_needed, showProgress = FALSE)
  setnames(dt, gsub("^off\\.", "a0.", names(dt)))

  row <- dt[replicate == rep & generation == gen]
  if (nrow(row) != 1) stop("Could not find exactly one row for replicate=", rep, " generation=", gen)

  out <- matrix(0, nrow = n_patches, ncol = 4)
  colnames(out) <- paste0("stage", 0:3)

  for (s in 0:3) {
    cols <- grep(paste0("^a", s, "\\.fem\\.p\\d+$"), names(row), value = TRUE)
    pnum <- as.integer(sub(paste0("^a", s, "\\.fem\\.p"), "", cols))
    ord  <- order(pnum)
    vals <- as.numeric(unlist(row[, ..cols])[ord])
    out[pnum[ord], s+1] <- vals
  }

  out
}

baseline_manifest <- data.table()

missing <- burnin_manifest[!file.exists(burnin_txt)]
if (nrow(missing) > 0) {
  warning("Missing burn-in outputs (run burn-ins first). Example missing file: ", missing$burnin_txt[1])
} else {
  for (i in seq_len(nrow(burnin_manifest))) {
    K <- burnin_manifest$K[i]
    B <- burnin_manifest$B[i]
    btag <- burnin_manifest$btag[i]
    txt <- burnin_manifest$burnin_txt[i]

    baseline_stage <- extract_baseline_stage_matrix(txt, n_patches, gen = BURNIN_GENS, rep = 1)
    stopifnot(nrow(baseline_stage) == n_patches, ncol(baseline_stage) == 4)

    out_file <- file.path(
      DIRS$stage,
      sprintf("patch_init_stage_size_baseline_k%d_b%s_gen%03d_rep%02d.txt", K, btag, BURNIN_GENS, 1)
    )
    write.matrix.nemo(baseline_stage, out_file)

    baseline_manifest <- rbind(
      baseline_manifest,
      data.table(K = K, B = B, btag = btag, baseline_stage_file = out_file)
    )
  }

  baseline_manifest_file <- file.path(OUT$scripts, "baseline_manifest.tsv")
  fwrite(baseline_manifest, baseline_manifest_file, sep = "\t")
  message("Wrote baseline stage files: ", nrow(baseline_manifest))
  message("Baseline manifest: ", baseline_manifest_file)
}
```

```{r}
#| label: layout-generators
#| echo: true

# utility: square layer distance (Chebyshev) around a center in (row,col)
rank_by_square_layers <- function(rows, cols, center_r, center_c, jitter = 0, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  d <- pmax(abs(rows - center_r), abs(cols - center_c))
  if (jitter > 0) d <- d + runif(length(d), -jitter, jitter)

  order(d)
}

# 1) dispersed: Poisson-disc-like sequential selection with min-distance relaxation
select_orientalis_dispersed <- function(
  nO,
  patch_ids,
  dist_mat,
  base_spacing_m,
  seed = NULL,
  strength = 1.20,
  relax_step = 0.95,
  max_relax = 12
) {
  if (!is.null(seed)) set.seed(seed)
  if (nO <= 0) return(integer(0))
  if (nO >= length(patch_ids)) return(patch_ids)

  minDist <- base_spacing_m * strength

  for (attempt in seq_len(max_relax)) {
    chosen <- integer(0)

    for (cand in sample(patch_ids)) {
      if (!length(chosen) || min(dist_mat[cand, chosen]) >= minDist) {
        chosen <- c(chosen, cand)
      }

      if (length(chosen) >= nO) return(chosen[1:nO])
    }

    minDist <- minDist * relax_step
  }

  # fallback
  sample(patch_ids, nO)
}

# 2) one square-ish cluster around (near) center, with slight edge softness via jitter
select_orientalis_one_cluster <- function(nO, coords_df, seed = NULL, jitter = 0.15) {
  if (!is.null(seed)) set.seed(seed)
  if (nO <= 0) return(integer(0))
  if (nO >= nrow(coords_df)) return(coords_df$patchID)

  # random small shift around center to create different layout runs
  center_r <- round(N / 2) + sample(-2:2, 1)
  center_c <- round(N / 2) + sample(-2:2, 1)

  center_r <- max(1, min(N, center_r))
  center_c <- max(1, min(N, center_c))

  ord <- rank_by_square_layers(
    coords_df$row,
    coords_df$col,
    center_r,
    center_c,
    jitter = jitter,
    seed = seed
  )

  coords_df$patchID[ord][1:nO]
}

# 3) multiple square-ish clusters with roughly equal size, placed roughly evenly across the grid
select_orientalis_multi_cluster <- function(
  nO,
  coords_df,
  seed = NULL,
  n_centers = 4,
  jitter = 0.15
) {
  if (!is.null(seed)) set.seed(seed)
  if (nO <= 0) return(integer(0))
  if (nO >= nrow(coords_df)) return(coords_df$patchID)

  n_centers <- max(2, min(n_centers, 6))

  # choose center locations by partitioning the grid into n_centers regions
  if (n_centers == 3) {
    centers_rc <- rbind(
      c(round(N * 0.25), round(N * 0.25)),
      c(round(N * 0.75), round(N * 0.25)),
      c(round(N * 0.50), round(N * 0.75))
    )
  } else { # default 4
    centers_rc <- rbind(
      c(round(N * 0.25), round(N * 0.25)),
      c(round(N * 0.75), round(N * 0.25)),
      c(round(N * 0.25), round(N * 0.75)),
      c(round(N * 0.75), round(N * 0.75))
    )

    if (n_centers > 4) {
      extra <- replicate(
        n_centers - 4,
        c(sample(5:(N - 4), 1), sample(5:(N - 4), 1))
      )
      centers_rc <- rbind(centers_rc, t(extra))
    }
  }

  # jitter centers slightly for each run
  centers_rc <- centers_rc +
    matrix(sample(-2:2, nrow(centers_rc) * 2, replace = TRUE), ncol = 2)

  centers_rc[, 1] <- pmin(N, pmax(1, centers_rc[, 1]))
  centers_rc[, 2] <- pmin(N, pmax(1, centers_rc[, 2]))

  # sizes per cluster (near-equal)
  base  <- floor(nO / nrow(centers_rc))
  sizes <- rep(base, nrow(centers_rc))
  sizes[1:(nO - sum(sizes))] <- sizes[1:(nO - sum(sizes))] + 1

  chosen    <- integer(0)
  available <- coords_df$patchID

  for (k in seq_len(nrow(centers_rc))) {
    center_r <- centers_rc[k, 1]
    center_c <- centers_rc[k, 2]

    sub <- coords_df[match(available, coords_df$patchID), ]

    ord <- rank_by_square_layers(
      sub$row,
      sub$col,
      center_r,
      center_c,
      jitter = jitter,
      seed = seed + k
    )

    take <- sub$patchID[ord][1:min(sizes[k], length(ord))]

    chosen    <- c(chosen, take)
    available <- setdiff(available, take)

    if (length(chosen) >= nO) break
  }

  chosen <- unique(chosen)

  if (length(chosen) > nO) {
    chosen <- chosen[1:nO]
  }

  if (length(chosen) < nO) {
    chosen <- c(
      chosen,
      sample(setdiff(coords_df$patchID, chosen), nO - length(chosen))
    )
  }

  chosen
}

# 4) transects (hard edges): full-length straight columns, approx matching p
select_orientalis_transects <- function(
  nO,
  coords_df,
  prop_orientalis = NULL,
  seed = NULL,
  ...
) {
  if (!is.null(seed)) set.seed(seed)
  if (nO <= 0) return(integer(0))
  if (nO >= nrow(coords_df)) return(coords_df$patchID)

  n_full <- nO %/% N
  rem    <- nO %% N

  full_cols <- if (n_full > 0) sample(1:N, n_full, replace = FALSE) else integer(0)

  chosen <- coords_df$patchID[coords_df$col %in% full_cols]

  if (rem > 0) {
    remaining_cols <- setdiff(1:N, full_cols)
    partial_col    <- sample(remaining_cols, 1)

    # continuous partial transect: fill from top OR bottom
    from_top    <- sample(c(TRUE, FALSE), 1)
    partial_rows <- if (from_top) 1:rem else (N - rem + 1):N

    chosen <- c(
      chosen,
      coords_df$patchID[
        coords_df$col == partial_col &
          coords_df$row %in% partial_rows
      ]
    )
  }

  chosen
}

# master wrapper
select_orientalis <- function(
  nO,
  configuration,
  patch_ids,
  dist_mat,
  coords_df,
  prop_orientalis,
  seed = NULL
) {
  configuration <- match.arg(configuration, CONFIG_LEVELS)

  if (configuration == "dispersed") {
    spacing_m <- sqrt(length(patch_ids) / max(1, nO)) * GRID_RES

    return(
      select_orientalis_dispersed(
        nO,
        patch_ids,
        dist_mat,
        base_spacing_m = spacing_m,
        seed = seed,
        strength = DISPERSED_STRENGTH
      )
    )
  }

  if (configuration == "one_cluster") {
    return(
      select_orientalis_one_cluster(
        nO,
        coords_df,
        seed = seed,
        jitter = CLUSTER_JITTER
      )
    )
  }

  if (configuration == "multi_cluster") {
    return(
      select_orientalis_multi_cluster(
        nO,
        coords_df,
        seed = seed,
        n_centers = MULTI_N_CLUSTERS,
        jitter = CLUSTER_JITTER
      )
    )
  }

  if (configuration == "transects") {
    return(
      select_orientalis_transects(
        nO,
        coords_df,
        prop_orientalis = prop_orientalis,
        seed = seed,
        min_tr = TRANSECTS_MIN,
        max_tr = TRANSECTS_MAX
      )
    )
  }

  stop("Unknown configuration: ", configuration)
}
```

## Generate scenario inputs

## Scenario writers + plotting helpers

```{r}
#| label: scenario-writers
#| echo: true

scenario_id <- function(configuration, prop_orientalis, run_id) {
  sprintf("%s_p%02d_r%02d", configuration, round(100 * prop_orientalis), run_id)
}

build_config_table <- function(orientalis_ids, patch_ids) {
  dt <- data.table(patchID = patch_ids, patch_value = "S")
  dt[patchID %in% orientalis_ids, patch_value := "O"]
  dt[order(patchID)]
}

write_config_quarto_schema <- function(conf_dt, sim_name, outfile) {
  out <- copy(conf_dt)
  out[, simulation := sim_name]
  setcolorder(out, c("simulation", "patchID", "patch_value"))
  fwrite(out, outfile, sep = "\t")
}

# quanti init (one column): 1 in O patches, 0 otherwise
build_quanti_init <- function(conf_dt, n_patches) {
  mat <- matrix(0, nrow = n_patches, ncol = 1)
  mat[conf_dt$patchID, 1] <- ifelse(conf_dt$patch_value == "O", 1, 0)
  mat
}

# patch_init_stage_size for run: start from baseline, clear O patches, plant orientalis scheme
build_patch_init_stage_size_from_baseline <- function(conf_dt, baseline_stage_mat, scheme = c(0,20,0,0)) {
  stopifnot(nrow(baseline_stage_mat) == nrow(conf_dt))
  out <- baseline_stage_mat
  O <- conf_dt$patchID[conf_dt$patch_value == "O"]
  out[O, ] <- 0
  out[O, ] <- matrix(rep(scheme, each = length(O)), nrow = length(O), byrow = FALSE)
  out
}

plot_grid_binary_gg <- function(conf_dt, coords_df, title = NULL) {
  df <- merge(coords_df, conf_dt, by = "patchID", all.x = TRUE)
  df$patch_value <- ifelse(is.na(df$patch_value), "S", df$patch_value)

  ggplot(df, aes(x = x_plot, y = y_plot, fill = patch_value)) +
    geom_tile(color = "grey85", linewidth = 0.1) +
    coord_equal() +
    labs(title = title, x = "", y = "", fill = "") +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      legend.position = "none",
      plot.title = element_text(size = 10, hjust = 0.5))
}
```

## Generate planting layouts (scenario configs + quanti init)

> Note: stage init files for scenarios are created *after* baseline extraction, per K×B.

```{r}
#| label: generate-scenarios
#| echo: true

manifest <- data.table()

for (p in PROP_LEVELS) {
  for (cfg in CONFIG_LEVELS) {
    for (r in seq_len(N_LAYOUT_RUNS)) {

      sim_name <- scenario_id(cfg, p, r)
      nO <- as.integer(round(p * n_patches))

      seed <- 10000 + round(1000*p) * 100 + r * 7 + nchar(cfg)  # deterministic-ish seed

      # --- choose Orientalis patches
      O_ids <- select_orientalis(
        nO = nO,
        configuration = cfg,
        patch_ids = patch_ids,
        dist_mat = dist_mat,
        coords_df = coords,
        prop_orientalis = p,
        seed = seed
      )

      # --- configs_txt (S/O per patch)
      cfg_dt <- build_config_table(O_ids, patch_ids)
      cfg_file <- file.path(DIRS$cfg, paste0(sim_name, ".txt"))
      write_config_quarto_schema(cfg_dt, sim_name, cfg_file)

      # --- run init quanti (scenario)
      q_mat <- build_quanti_init(cfg_dt, n_patches)
      q_file <- file.path(DIRS$quanti, paste0("quanti_init_file_", sim_name, ".txt"))
      write.matrix.nemo(q_mat, q_file)

      manifest <- rbind(
        manifest,
        data.table(
          sim_name = sim_name,
          configuration = cfg,
          prop_orientalis = p,
          n_orientalis = nO,
          run_id = r,
          seed = seed,
          cfg_file = cfg_file,
          quanti_file = q_file
        )
      )
    }
  }
}

manifest_file <- file.path(OUT$scripts, "layout_manifest.tsv")
fwrite(manifest, manifest_file, sep = "\t")
message("Generated ", nrow(manifest), " layouts.")
message("Layout manifest: ", manifest_file)
```

## Layout plots

```{r}
#| label: scenario-plots
#| echo: true

PLOT_DIR <- file.path(DIRS$plots, "layout_runs")
dir.create(PLOT_DIR, showWarnings = FALSE, recursive = TRUE)

manifest_dt <- copy(manifest)
manifest_dt[, scenario_key := sub("_r[0-9]+$", "", sim_name)]  # group by config × p

scenario_groups <- manifest_dt[
  ,
  .(
    cfg_files = list(cfg_file),
    sims = list(sim_name),
    prop = prop_orientalis[1],
    configuration = configuration[1],
    n_orientalis = n_orientalis[1]
  ),
  by = .(scenario_key)
]

for (i in seq_len(nrow(scenario_groups))) {
  key <- scenario_groups$scenario_key[i]
  cfg <- scenario_groups$configuration[i]
  p   <- scenario_groups$prop[i]

  cfg_files <- unlist(scenario_groups$cfg_files[i])
  sims      <- unlist(scenario_groups$sims[i])

  ord <- order(sims)
  cfg_files <- cfg_files[ord]
  sims <- sims[ord]
  
  plist <- lapply(seq_along(cfg_files), function(j) {
    dt <- fread(cfg_files[j])[, .(patchID, patch_value)]
    nO_run <- sum(dt$patch_value == "O")
    rep_lab <- sprintf("r%02d | %d (%.2f%%)", j, nO_run, 100 * nO_run / n_patches)
    plot_grid_binary_gg(dt, coords, rep_lab)
  })

  nO_used <- scenario_groups$n_orientalis[i]
  main_title <- sprintf(
    "%s | requested %.0f%% (%.1f/%d) | used %d/%d = %.2f%%",
    cfg,
    100 * p,
    p * n_patches,
  n_patches,
  nO_used,
  n_patches,
  100 * nO_used / n_patches
)


  panel <- cowplot::plot_grid(plotlist = plist, nrow = 1, align = "hv")
  final <- cowplot::plot_grid(
    cowplot::ggdraw() + cowplot::draw_label(main_title, fontface = "bold", size = 14),
    panel,
    ncol = 1,
    rel_heights = c(0.18, 1)
  )
  
  final <- final + theme(plot.background = element_rect(fill = "white", color = NA))

  out_png <- file.path(PLOT_DIR, paste0("overview_", key, ".png"))
  ggsave(out_png, final, width = 16, height = 3.8, dpi = 220)
}

message("Saved layout overviews in: ", PLOT_DIR)
```

## Write run INIs for all scenarios × K×B

```{r}
#| label: write-run-inis
#| echo: true

if (!exists("baseline_manifest") || nrow(baseline_manifest) == 0) {
  stop("No baseline_manifest found. Run burn-ins, then re-run this document to build baselines.")
}

run_template <- read_ini_lines(TEMPLATE_INI)

# fixed parts
run_template <- set_ini_param(run_template, "run_mode", "overwrite")
run_template <- set_ini_param(run_template, "replicates", NEMO_REPS_RUN)
run_template <- set_ini_param(run_template, "generations", RUN_GENS)

# logging
run_template <- set_ini_param(run_template, "stat_log_time",
                              paste0("{{", paste(LOG_TIMES, collapse = ","), "}}"))
run_template <- set_ini_param(run_template, "quanti_dir", "quanti")
run_template <- set_ini_param(run_template, "quanti_logtime",
                              paste0("{{", paste(LOG_TIMES, collapse = ","), "}}"))

# dispersal
run_template <- set_ini_param(run_template, "seed_disperse_connectivity_matrix", paste0("&", seed_conn_file))
run_template <- set_ini_param(run_template, "seed_disperse_reduced_matrix",       paste0("&", seed_rate_file))
run_template <- set_ini_param(run_template, "breed_disperse_connectivity_matrix_mal", paste0("&", pollen_conn_file))
run_template <- set_ini_param(run_template, "breed_disperse_reduced_matrix_mal",       paste0("&", pollen_rate_file))

# Stage file builder: baseline + clear O patches + plant Orientalis scheme
build_stage_from_baseline <- function(baseline_mat, cfg_dt, scheme) {
  out <- baseline_mat
  O_ids <- cfg_dt$patchID[cfg_dt$patch_value == "O"]
  out[O_ids, ] <- 0
  out[O_ids, ] <- matrix(rep(scheme, each = length(O_ids)), nrow = length(O_ids), byrow = FALSE)
  out
}

# Ensure scenario stage subfolder
stage_supp_dir <- file.path(DIRS$stage, "scenarios")
dir.create(stage_supp_dir, showWarnings = FALSE, recursive = TRUE)

ini_manifest <- data.table()

for (i in seq_len(nrow(manifest))) {
  sim  <- manifest$sim_name[i]
  cfg  <- manifest$configuration[i]
  p    <- manifest$prop_orientalis[i]
  rid  <- manifest$run_id[i]
  seed <- manifest$seed[i]

  # read cfg_dt to know O patches
  cfg_dt <- fread(manifest$cfg_file[i])
  cfg_dt[, patchID := as.integer(patchID)]

  for (j in seq_len(nrow(baseline_manifest))) {
    K <- baseline_manifest$K[j]
    B <- baseline_manifest$B[j]
    btag <- baseline_manifest$btag[j]
    base_file <- baseline_manifest$baseline_stage_file[j]

    baseline_mat <- read.matrix.nemo(base_file, nrow_expected = n_patches, ncol_expected = 4)

    stage_mat <- build_stage_from_baseline(baseline_mat, cfg_dt, PLANTING_SCHEME)

    stage_file <- file.path(stage_supp_dir, sprintf("patch_init_stage_size_%s_k%d_b%s.txt", sim, K, btag))
    write.matrix.nemo(stage_mat, stage_file)

    # directory grouping for results
    run_root <- file.path(
      OUT$results, cfg,
      sprintf("p%02d", round(100*p)),
      sprintf("r%02d", rid),
      sprintf("k%d_b%s", K, btag),
      "run"
    )
    dir.create(run_root, showWarnings = FALSE, recursive = TRUE)

    ini <- run_template
    ini <- set_ini_param(ini, "random_seed", seed + 999 + K * 10 + as.integer(round(B * 1e4)))
    ini <- set_ini_param(ini, "root_dir", run_root)
    ini <- set_ini_param(ini, "filename", sprintf("%s_k%d_b%s", sim, K, btag))
    ini <- set_ini_param(ini, "logfile",  sprintf("%s_k%d_b%s_log", sim, K, btag))

    ini <- set_ini_param(ini, "patch_nbfem", K)
    ini <- set_ini_param(ini, "regulation_by_competition", B)

    ini <- set_ini_param(ini, "quanti_init_freq", paste0("&", manifest$quanti_file[i]))
    ini <- set_ini_param(ini, "patch_init_stage_size", paste0("&", stage_file))

    ini_path <- file.path(DIRS$ini_run, sprintf("%s_k%d_b%s.ini", sim, K, btag))
    writeLines(ini, ini_path)

    ini_manifest <- rbind(
      ini_manifest,
      data.table(
        sim_name = sim,
        configuration = cfg,
        prop_orientalis = p,
        run_id = rid,
        K = K,
        B = B,
        btag = btag,
        baseline_stage_file = base_file,
        stage_file = stage_file,
        ini = ini_path,
        run_root = run_root
      )
    )
  }
}

ini_manifest_file <- file.path(OUT$scripts, "ini_manifest.tsv")
fwrite(ini_manifest, ini_manifest_file, sep = "\t")
message("Wrote run INIs: ", nrow(ini_manifest))
message("INI manifest: ", ini_manifest_file)
```

## Script to run all scenario simulations

```{r}
#| label: write-run-script
#| echo: true

cmds <- sprintf("./nemoage %s", ini_manifest$ini)

sh_all <- file.path(OUT$scripts, "run_all_runs.sh")
writeLines(c("#!/usr/bin/env bash", "set -euo pipefail", "", cmds), sh_all)
Sys.chmod(sh_all, mode = "0755")

message("Wrote: ", sh_all)
```

Run scenarios (after baselines + INIs exist):

``` bash
./supplementary/scripts/run_all_runs.sh
```
