---
title: "Run Diagnostics: Demography + Hybrid Distributions"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
params:
  project_root: "~/Documents/WSL/WSL-sylvatica-orientalis-hybridization"
  configuration: "dispersed"
  use_cache: false
  layout_run: 1
  p_focus: 25
  gen_max: 500
  gen_keep_every: 50
  max_internal_replicate: 20
  quanti_stage: 3
  p1_bin_width: 0.1
  p1_tol: 1e-8
  out_dir: "run_outputs"
  save_plots: true
---

## Setup

```{r}
library(data.table)
library(stringr)
library(dplyr)
library(ggplot2)
library(scales)
library(viridis)

project_root <- path.expand(params$project_root)
res_root     <- file.path(project_root, "results")

out_root  <- file.path(project_root, params$out_dir)
plot_dir  <- file.path(out_root, "plots")
cache_dir <- file.path(out_root, "cache")

dir.create(plot_dir,  recursive = TRUE, showWarnings = FALSE)
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

save_plot <- function(p, name, w = 14, h = 8, dpi = 220) {
  if (!isTRUE(params$save_plots)) return(invisible(NULL))
  fn <- file.path(plot_dir, paste0(name, ".png"))
  ggsave(fn, p, width = w, height = h, dpi = dpi, bg = "white")
  invisible(fn)
}
```

### Filename metadata parser

Examples: - `dispersed_p10_r01_k20_b0.01.txt` - `dispersed_p10_r01_k20_b0.01_050_01.quanti`

```{r}
parse_meta_from_basename <- function(path) {
  bname <- basename(path)
  m <- str_match(
    bname,
    "^(.+?)_p(\\d+)_r(\\d+)_k(\\d+(?:\\.\\d+)?)_b(\\d+(?:\\.\\d+)?)(?:_|\\.|$)"
  )
  if (any(is.na(m))) stop("Could not parse metadata from filename: ", bname)
  data.table(
    configuration  = m[,2],
    ori_proportion = as.integer(m[,3]),
    layout_run     = as.integer(m[,4]),
    k              = as.numeric(m[,5]),
    b              = as.numeric(m[,6])
  )
}

parse_quanti_suffix <- function(path) {
  bname <- basename(path)
  # accept 1–3 digits for generation token, e.g. _2_01.quanti or _002_01.quanti or _500_01.quanti
  m <- str_match(bname, "_(\\d{1,3})_(\\d+)\\.quanti$")
  if (any(is.na(m))) stop("Could not parse quanti suffix: ", bname)
  list(generation = as.integer(m[,2]), replicate = as.integer(m[,3]))
}

run_str <- sprintf("r%02d", as.integer(params$layout_run))
cfg     <- params$configuration
```

## File discovery (main run only)

We only take files under: `results/dispersed/pXX/r01/run/...`

```{r}
# --- demography .txt (run summaries; exclude *_bygen.txt)
demog_files <- list.files(file.path(res_root, cfg),
                          recursive = TRUE, full.names = TRUE, pattern = "\\.txt$")

demog_files <- demog_files[grepl(paste0("/", cfg, "/p\\d+/", run_str, "/run/"), gsub("\\\\","/", demog_files))]
demog_files <- demog_files[grepl(paste0("^", cfg, "_p\\d+_", run_str, "_k[0-9.]+_b[0-9.]+\\.txt$"), basename(demog_files))]
demog_files <- demog_files[!grepl("_bygen\\.txt$", demog_files)]
demog_files <- demog_files[!grepl("\\.log$", demog_files)]

# --- quanti .quanti (accept quanti/qunati)
quanti_files <- list.files(file.path(res_root, cfg),
                           recursive = TRUE, full.names = TRUE, pattern = "\\.quanti$")

quanti_files <- quanti_files[grepl(paste0("/", cfg, "/p\\d+/", run_str, "/run/(quanti|qunati)/"), gsub("\\\\","/", quanti_files))]
quanti_files <- quanti_files[grepl(paste0("^", cfg, "_p\\d+_", run_str, "_k[0-9.]+_b[0-9.]+_\\d{1,3}_\\d+\\.quanti$"),
                                   basename(quanti_files))]

length(demog_files)
length(quanti_files)
```

------------------------------------------------------------------------

# Demography: replicate cloud + mean

## Read and cache demography (stand totals)

```{r}
# tag for filenames (so pall vs p25 etc)
p_tag <- if (is.null(params$p_focus)) "pall" else paste0("p", as.integer(params$p_focus))

demog_rds <- file.path(
  cache_dir,
  paste0("run_demog_long_", cfg, "_", run_str, "_", p_tag, "_gmax", params$gen_max, ".rds")
)

cols_to_read <- c("replicate", "generation", "pop.tot", "a0.tot", "a1.tot", "a2.tot", "a3.tot")

if (isTRUE(params$use_cache) && file.exists(demog_rds)) {
  demog_long <- readRDS(demog_rds)
} else {
  dl <- vector("list", length(demog_files))

  for (i in seq_along(demog_files)) {
    f <- demog_files[i]
    meta <- parse_meta_from_basename(f)

    header <- names(fread(f, nrows = 0))
    keep <- intersect(cols_to_read, header)
    dt <- fread(f, select = keep)

    dt <- cbind(dt, meta)

    dl[[i]] <- melt(
      dt,
      id.vars = c("replicate","generation","configuration","ori_proportion","layout_run","k","b"),
      measure.vars = patterns("\\.tot$"),
      variable.name = "stage",
      value.name = "N"
    )
  }

  demog_long <- rbindlist(dl, use.names = TRUE, fill = TRUE)

  # filters: internal replicates + time window
  demog_long <- demog_long[replicate %between% c(1, params$max_internal_replicate)]
  demog_long <- demog_long[generation <= params$gen_max]

  # filter by p if requested (THIS is what you were missing)
  if (!is.null(params$p_focus)) {
    demog_long <- demog_long[ori_proportion == as.integer(params$p_focus)]
  }

  if (isTRUE(params$use_cache)) saveRDS(demog_long, demog_rds)
}

stage_labs <- c(
  "pop.tot" = "Total population",
  "a0.tot"  = "Stage 0",
  "a1.tot"  = "Stage 1",
  "a2.tot"  = "Stage 2",
  "a3.tot"  = "Stage 3"
)
demog_long[, stage_label := stage_labs[stage]]
demog_long[, kb := paste0("k", k, " b", b)]
```

```{r}
demog_mean <- demog_long[
  , .(meanN = mean(N, na.rm = TRUE)),
  by = .(generation, stage_label, ori_proportion, k, b)
]
demog_mean[, kb := paste0("k", k, " b", b)]
```

## Plot: cloud (all stages)

```{r}

p_dem_cloud <- ggplot() +
  geom_line(
    data = demog_long,
    aes(x = generation, y = N,
        group = interaction(stage_label, replicate),
        color = stage_label),
    linewidth = 0.15, alpha = 0.4
  ) +
  geom_line(
    data = demog_mean,
    aes(x = generation, y = meanN, color = stage_label),
    linewidth = 0.7
  ) +
  facet_grid(ori_proportion ~ kb, scales = "free_y") +
  labs(
    title = "Burn-in demography: replicate variability (faint) + mean (bold)",
    x = "Generation", y = "Abundance", color = "Stage"
  ) +
  scale_y_continuous(labels = comma) +
  theme_bw()

p_dem_cloud
save_plot(p_dem_cloud, paste0("run_demography_cloud_all_stages_", params$p_focus, "_supplementary"), w = 20, h = 10)
```

## Plot: cloud (adults only)

```{r}
dem_ad_long <- demog_long[stage_label == "Stage 3"]
dem_ad_mean <- demog_mean[stage_label == "Stage 3"]

p_dem_cloud_adults <- ggplot() +
  geom_line(
    data = dem_ad_long,
    aes(x = generation, y = N, group = replicate),
    linewidth = 0.35, alpha = 0.35, color = "dodgerblue"
  ) +
  geom_line(
    data = dem_ad_mean,
    aes(x = generation, y = meanN),
    linewidth = 1.25, color = "dodgerblue"
  ) +
  facet_grid(ori_proportion ~ kb, scales = "free_y") +
  labs(
    title = "Adult demography cloud (replicates) + mean (bold)",
    x = "Generation", y = "Adult abundance (Stage 3)"
  ) +
  scale_y_continuous(labels = comma) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "grey98", colour = NA),
    panel.grid.major = element_line(colour = "grey88", linewidth = 0.3),
    panel.grid.minor = element_line(colour = "grey92", linewidth = 0.2)
  )

p_dem_cloud_adults
save_plot(p_dem_cloud_adults, paste0("run_demography_cloud_adults_only_", params$p_focus, "_supplementary"), w = 20, h = 10)
```

------------------------------------------------------------------------

# Quanti (P1): distribution over time and hybrids

```{r}
p1_bins_rds    <- file.path(cache_dir, paste0("run_quanti_p1_bins_", cfg, "_", run_str, "_gmax", params$gen_max, ".rds"))
p1_metrics_rds <- file.path(cache_dir, paste0("run_quanti_p1_metrics_", cfg, "_", run_str, "_gmax", params$gen_max, ".rds"))

# bins for P1
bw  <- as.numeric(params$p1_bin_width)
breaks <- seq(-1, 1, by = bw)
if (tail(breaks, 1) < 1) breaks <- c(breaks, 1)

tol <- as.numeric(params$p1_tol)
STAGE <- as.integer(params$quanti_stage)

build_quanti_summaries <- function() {
  bin_list <- list()
  met_list <- list()

  for (f in quanti_files) {
    meta <- parse_meta_from_basename(f)
    suf  <- parse_quanti_suffix(f)

    if (suf$generation > params$gen_max) next

    dt <- fread(f, select = c("stage", "P1"))
    dt <- dt[stage == STAGE]
    if (nrow(dt) == 0) next

    # classify
    dt[, is_sylv := abs(P1 + 1) <= tol]
    dt[, is_ori  := abs(P1 - 1) <= tol]
    dt[, is_f1   := abs(P1) <= tol]
    dt[, is_intro := (P1 > (-1 + tol)) & (P1 < (1 - tol))]

    N <- nrow(dt)

    met_list[[length(met_list) + 1]] <- cbind(
      data.table(
        generation = suf$generation,
        replicate  = suf$replicate,
        N = N,
        prop_f1 = mean(dt$is_f1, na.rm = TRUE),
        prop_intro = mean(dt$is_intro, na.rm = TRUE),
        mean_P1 = mean(dt$P1, na.rm = TRUE)
      ),
      meta
    )

    # make sure P1 is numeric
    dt[, P1 := as.numeric(P1)]

    # assign each P1 to a numeric bin id (1..nbins)
    dt[, bin_id := findInterval(P1, breaks, rightmost.closed = TRUE, all.inside = TRUE)]

    # compute bin midpoints directly from breaks
    bin_mids <- (breaks[-length(breaks)] + breaks[-1]) / 2

    # attach midpoint to each row
    dt[, bin_mid := bin_mids[bin_id]]

    # summarize by midpoint
    bins <- dt[, .N, by = .(bin_mid)]
    bins[, prop := N / sum(N)]

    bins[, `:=`(generation = suf$generation, replicate = suf$replicate)]
    bins <- cbind(bins, meta)

    bin_list[[length(bin_list) + 1]] <- bins
  }

  list(
    p1_bins = rbindlist(bin_list, use.names = TRUE, fill = TRUE),
    p1_metrics = rbindlist(met_list, use.names = TRUE, fill = TRUE)
  )
}

if (isTRUE(params$use_cache) && file.exists(p1_bins_rds) && file.exists(p1_metrics_rds)) {
  p1_bins <- readRDS(p1_bins_rds)
  p1_metrics <- readRDS(p1_metrics_rds)
} else {
  tmp <- build_quanti_summaries()
  p1_bins <- tmp$p1_bins
  p1_metrics <- tmp$p1_metrics

  if (isTRUE(params$use_cache)) {
    saveRDS(p1_bins, p1_bins_rds)
    saveRDS(p1_metrics, p1_metrics_rds)
  }
}


# enforce internal replicate range + optional p focus
p1_bins    <- p1_bins[replicate %between% c(1, params$max_internal_replicate)]
p1_metrics <- p1_metrics[replicate %between% c(1, params$max_internal_replicate)]

if (!is.null(params$p_focus)) {
  p1_bins    <- p1_bins[ori_proportion == as.integer(params$p_focus)]
  p1_metrics <- p1_metrics[ori_proportion == as.integer(params$p_focus)]
}

p1_bins[, kb := paste0("k", k, " b", b)]
p1_metrics[, kb := paste0("k", k, " b", b)]
```

## Plot: P1 distribution over time (tile plot; adults)

Each tile shows the **mean proportion** (across internal replicates) of individuals whose P1 falls into that bin at that generation.

```{r}
# Ensure bins missing in a replicate are treated as 0
# Use expected bin midpoints from breaks (not just those observed)
bin_mids_expected <- (breaks[-length(breaks)] + breaks[-1]) / 2
bin_mids_expected <- round(bin_mids_expected, 2)

# Round bin_mid in p1_bins too (prevents floating-point join mismatch)
p1_bins[, bin_mid := round(as.numeric(bin_mid), 2)]

# Build grid: all scenario×generation×replicate keys × all expected bins
keys <- unique(p1_bins[, .(configuration, ori_proportion, layout_run, k, b, kb, generation, replicate)])

grid <- keys[
  , .(bin_mid = bin_mids_expected),
  by = .(configuration, ori_proportion, layout_run, k, b, kb, generation, replicate)
]

# Left join observed bins onto the full grid, then fill missing with 0
p1_bins_full <- merge(
  grid, p1_bins,
  by = c("configuration","ori_proportion","layout_run","k","b","kb","generation","replicate","bin_mid"),
  all.x = TRUE
)
p1_bins_full[is.na(prop), prop := 0]

# Mean across internal replicates
p1_bin_mean <- p1_bins_full[
  , .(mean_prop = mean(prop)),
  by = .(generation, bin_mid, ori_proportion, k, b, kb)
]

p1_bin_mean[, gen_f := factor(generation, levels = sort(unique(generation)))]

p_p1_tile <- ggplot(p1_bin_mean, aes(x = gen_f, y = bin_mid, fill = mean_prop)) +
  geom_tile() +
  facet_grid(ori_proportion ~ kb) +
  scale_fill_viridis_c(option = "C", trans = "sqrt") +
  labs(
    title = paste0("P1 distribution over time (stage ", STAGE, "; mean across internal replicates)"),
    x = "Generation",
    y = "P1 bin midpoint (-1 = sylvatica, 0 = hybrid, +1 = orientalis)",
    fill = "Mean\nproportion"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white", colour = NA)
  )

p_p1_tile
save_plot(p_p1_tile, paste0("run_quanti_P1_distribution_tile_", params$p_focus, "_supplementary"), w = 20, h = 10)

```

## Plot: How many hybrids over time (F1 = P1=0)

```{r}
p1_f1_time <- p1_metrics[
  , .(
    mean_f1 = mean(prop_f1, na.rm = TRUE)
  ),
  by = .(generation, ori_proportion, k, b, kb)
]

p_f1 <- ggplot(p1_f1_time, aes(x = generation, y = mean_f1)) +
  geom_line(linewidth = 0.8) +
  facet_grid(ori_proportion ~ kb) +
  labs(
    title = "F1 hybrid proportion over time (P1=0)",
    x = "Generation",
    y = "Proportion F1 (P1=0)"
  ) +
  theme_bw()

p_f1
save_plot(p_f1, paste0("run_quanti_F1_prop_time_series_", params$p_focus, "_supplementary"), w = 20, h = 10)
```

```{r}

library(ggplot2)
library(scales)
library(dplyr)
library(patchwork)

# ---- Filter out total pop
subsetdemog_long <- demog_long %>% filter(stage_label != "Total population")
subsetdemog_mean <- demog_mean %>% filter(stage_label != "Total population")

# ---- Nice labels for facets
lab_k <- function(x) paste0("K = ", x)
lab_b <- function(x) paste0("B = ", x)

# ---- Consistent stage order (optional but helps legend/order)
stage_levels <- c("Stage 0", "Stage 1", "Stage 2", "Stage 3")
subsetdemog_long$stage_label <- factor(subsetdemog_long$stage_label, levels = stage_levels)
subsetdemog_mean$stage_label <- factor(subsetdemog_mean$stage_label, levels = stage_levels)


# Panel A: all stages (cloud + mean), sqrt y to show structure

p_all <- ggplot() +
  geom_line(
    data = subsetdemog_long,
    aes(x = generation, y = N,
        group = interaction(stage_label, replicate),
        color = stage_label),
    linewidth = 0.18, alpha = 0.20
  ) +
  geom_line(
    data = subsetdemog_mean,
    aes(x = generation, y = meanN, color = stage_label),
    linewidth = 0.85
  ) +
  facet_grid(
    k ~ b,
    labeller = labeller(k = lab_k, b = lab_b),
    scales = "fixed",
  ) +
  scale_y_sqrt(
    breaks = c(0, 100, 500, 1000, 2500, 5000, 10000),
    labels = comma
  ) +
  labs(
    title = "Burn-in demography across parameter combinations",
    subtitle = "Faint lines = NEMO replicates, bold lines = mean trajectory",
    x = "Generation (years)",
    y = "Abundance (sqrt scale)",
    color = "Life stage"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey95", color = NA),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )


# Panel B: Adults only (Stage 3), linear y with target band

adult_long <- subsetdemog_long %>% filter(stage_label == "Stage 3")
adult_mean <- subsetdemog_mean %>% filter(stage_label == "Stage 3")

facet_keys <- expand.grid(
  k = sort(unique(adult_long$k)),
  b = sort(unique(adult_long$b))
)

p_adults <- ggplot() +
  geom_rect(
    data = facet_keys,
    aes(xmin = -Inf, xmax = Inf, ymin = 40, ymax = 60),
    inherit.aes = FALSE,
    alpha = 0.12
  ) +
  geom_hline(
    data = facet_keys,
    aes(yintercept = 50),
    inherit.aes = FALSE,
    linewidth = 0.4,
    linetype = "dashed"
  ) +
  geom_line(
    data = adult_long,
    aes(x = generation, y = N, group = replicate),
    linewidth = 0.25, alpha = 0.25
  ) +
  geom_line(
    data = adult_mean,
    aes(x = generation, y = meanN),
    linewidth = 1.0
  ) +
  facet_grid(k ~ b, scales = "fixed") +
  scale_y_continuous(breaks = pretty_breaks(n = 6), labels = comma) +
  labs(
    title = "Adults only (Stage 3): target ~50 adults",
    subtitle = "Red band = 40–60 adults; dashed line = 50",
    x = "Generation (years)",
    y = "Adults (linear scale)"
  ) +
  theme_bw()

p_adults

# Combine into one clean figure

final_demog_plot <- p_all / p_adults + plot_layout(heights = c(2.2, 1.6))

final_demog_plot

















library(ggplot2)
library(scales)
library(dplyr)
library(patchwork)

# -----------------------------
# Settings (palette, labels, theme)
# -----------------------------
stage_levels <- c("Stage 0", "Stage 1", "Stage 2", "Stage 3")

# Bright, clean, colorblind-friendly palette (change hex if you prefer)
stage_pal <- c(
  "Stage 0" = "#ff7f00",  # teal
  "Stage 1" = "#984ea3",  # orange
  "Stage 2" = "#4daf4a",  # purple
  "Stage 3" = "#377eb8"   # magenta (used in BOTH panels)
)

lab_k <- function(x) paste0("K = ", x)
lab_b <- function(x) paste0("B = ", x)

theme_demog <- theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.background  = element_rect(fill = "grey95", colour = NA),
    strip.text        = element_text(face = "bold"),
    legend.position   = "bottom",
    legend.title      = element_text(face = "bold"),
    plot.title        = element_text(face = "bold"),
    plot.subtitle     = element_text(margin = margin(b = 6)),
    plot.margin       = margin(8, 10, 8, 10)
  )

# -----------------------------
# Prep data
# -----------------------------
demog_long2 <- demog_long %>%
  filter(stage_label != "Total population") %>%
  mutate(stage_label = factor(stage_label, levels = stage_levels))

demog_mean2 <- demog_mean %>%
  filter(stage_label != "Total population") %>%
  mutate(stage_label = factor(stage_label, levels = stage_levels))

# Panel A: all stages (replicate cloud + mean), sqrt y

p_all <- ggplot(demog_long2, aes(generation, N, colour = stage_label)) +
  # Replicates: thin + transparent, but NOT in legend
  geom_line(
    aes(group = interaction(stage_label, replicate)),
    linewidth = 0.2, alpha = 0.6,
    show.legend = FALSE
  ) +
  # Means: thicker, shown in legend
  geom_line(
    data = demog_mean2,
    aes(y = meanN, group = stage_label),
    linewidth = 0.5
  ) +
  facet_grid(
    k ~ b,
    scales = "free_y",
    labeller = labeller(k = lab_k, b = lab_b)
  ) +
  scale_color_manual(values = stage_pal, drop = FALSE) +
  scale_x_continuous(
    breaks = pretty_breaks(n = 5),
    expand = expansion(mult = c(0.01, 0.02))
  ) +
  labs(
    title = "Burn-in demography across parameter combinations",
    subtitle = "Thin lines = replicates; thick lines = mean",
    x = "Generation (years)",
    y = "Abundance (sqrt scale)",
    colour = "Life stage"
  ) +
  guides(colour = guide_legend(override.aes = list(linewidth = 1.6, alpha = 1))) +
  theme_demog


# Panel B: Adults only (Stage 3), linear y with colored target band

adult_long <- demog_long2 %>% filter(stage_label == "Stage 3")
adult_mean <- demog_mean2 %>% filter(stage_label == "Stage 3")
facet_keys <- adult_long %>% distinct(k, b)

stage3_col <- stage_pal["Stage 3"]

p_adults <- ggplot(adult_long, aes(generation, N)) +
  # Colored band 40–60
  geom_rect(
    data = facet_keys,
    aes(xmin = -Inf, xmax = Inf, ymin = 40, ymax = 60),
    inherit.aes = FALSE,
    fill = "#e41a1c",
    alpha = 0.3,
    colour = NA
  ) +
  # Target line at 50
  geom_hline(
    data = facet_keys,
    aes(yintercept = 50),
    inherit.aes = FALSE,
    linewidth = 0.25,
    linetype = "dashed"
  ) +
  # Replicates (slightly tinted Stage 3)
  geom_line(
    aes(group = replicate),
    linewidth = 0.2,
    alpha = 0.6,
    colour = stage3_col
  ) +
  # Mean (strong Stage 3)
  geom_line(
    data = adult_mean,
    aes(y = meanN),
    linewidth = 0.5,
    colour = stage3_col
  ) +
  facet_grid(
    k ~ b,
    labeller = labeller(k = lab_k, b = lab_b)
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(n = 4),
    labels = comma,
    expand = expansion(mult = c(0.02, 0.06))
  ) +
  scale_x_continuous(
    breaks = pretty_breaks(n = 5),
    expand = expansion(mult = c(0.01, 0.02))
  ) +
  labs(
    title = "Adults only (Stage 3): target ≈ 50",
    subtitle = "Shaded band = 40–60; dashed line = 50",
    x = "Generation (years)",
    y = "Adults (linear scale)"
  ) +
  theme_demog +
  theme(legend.position = "none")


# Combine into one

final_demog_plot <- (p_all / p_adults) +
  plot_layout(heights = c(2.2, 1.6)) +
  plot_annotation(tag_levels = "A")

final_demog_plot

```
