---
title: "Run Diagnostics: Demography + Hybrid Distributions (All scenarios × all proportions)"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
params:
  project_root: "~/Documents/WSL/WSL-sylvatica-orientalis-hybridization"
  use_cache: false
  layout_run: 1
  gen_max: 500
  gen_keep_every: 50
  max_internal_replicate: 20
  quanti_stage: 3
  p1_bin_width: 0.1
  p1_tol: 1e-8
  out_dir: "run_outputs"
  save_plots: true
  configurations: ["dispersed","one_cluster","multi_cluster","transects"]
  proportions: [10,25,40]
---

## Setup

```{r}
library(data.table)
library(stringr)
library(dplyr)
library(ggplot2)
library(scales)
library(viridis)

project_root <- path.expand(params$project_root)
res_root     <- file.path(project_root, "results")

out_root  <- file.path(project_root, params$out_dir)
plot_dir  <- file.path(out_root, "plots")
cache_dir <- file.path(out_root, "cache")

dir.create(plot_dir,  recursive = TRUE, showWarnings = FALSE)
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

CFGs <- as.character(params$configurations)
PCTs <- as.integer(params$proportions)
run_str <- sprintf("r%02d", as.integer(params$layout_run))

safe_slug <- function(x) {
  x <- gsub("[^A-Za-z0-9]+", "_", x)
  x <- gsub("_+", "_", x)
  x <- gsub("^_|_$", "", x)
  x
}

save_plot <- function(p, name, w = 14, h = 8, dpi = 220) {
  if (!isTRUE(params$save_plots)) return(invisible(NULL))
  fn <- file.path(plot_dir, paste0(name, ".png"))
  ggsave(fn, p, width = w, height = h, dpi = dpi, bg = "white")
  invisible(fn)
}
```

### Filename metadata parser

Examples: - `dispersed_p10_r01_k20_b0.01.txt` - `dispersed_p10_r01_k20_b0.01_050_01.quanti`

```{r}
parse_meta_from_basename <- function(path) {
  bname <- basename(path)
  m <- str_match(
    bname,
    "^(.+?)_p(\\d+)_r(\\d+)_k(\\d+(?:\\.\\d+)?)_b(\\d+(?:\\.\\d+)?)(?:_|\\.|$)"
  )
  if (any(is.na(m))) stop("Could not parse metadata from filename: ", bname)
  data.table(
    configuration  = m[,2],
    ori_proportion = as.integer(m[,3]),
    layout_run     = as.integer(m[,4]),
    k              = as.numeric(m[,5]),
    b              = as.numeric(m[,6])
  )
}

parse_quanti_suffix <- function(path) {
  bname <- basename(path)
  m <- str_match(bname, "_(\\d{1,3})_(\\d+)\\.quanti$")
  if (any(is.na(m))) stop("Could not parse quanti suffix: ", bname)
  list(generation = as.integer(m[,2]), replicate = as.integer(m[,3]))
}
```

## File discovery (main run only)

We only take files under: `results/dispersed/pXX/r01/run/...`

```{r}
cfg_rx <- paste(CFGs, collapse = "|")

# demography .txt (exclude *_bygen.txt)
demog_files <- list.files(res_root, recursive = TRUE, full.names = TRUE, pattern = "\\.txt$")
demog_files <- demog_files[!grepl("_bygen\\.txt$", demog_files)]
demog_files <- demog_files[!grepl("\\.log$", demog_files)]

# keep only: results/<cfg>/pXX/rYY/run/...
demog_files <- demog_files[
  grepl(paste0("/(", cfg_rx, ")/p\\d+/", run_str, "/run/"), gsub("\\\\","/", demog_files))
]

# and enforce basename naming convention
demog_files <- demog_files[
  grepl(paste0("^(", cfg_rx, ")_p\\d+_", run_str, "_k[0-9.]+_b[0-9.]+\\.txt$"),
        basename(demog_files))
]

# quanti .quanti (accept quanti/qunati directories)
quanti_files <- list.files(res_root, recursive = TRUE, full.names = TRUE, pattern = "\\.quanti$")
quanti_files <- quanti_files[
  grepl(paste0("/(", cfg_rx, ")/p\\d+/", run_str, "/run/(quanti|qunati)/"), gsub("\\\\","/", quanti_files))
]
quanti_files <- quanti_files[
  grepl(paste0("^(", cfg_rx, ")_p\\d+_", run_str, "_k[0-9.]+_b[0-9.]+_\\d{1,3}_\\d+\\.quanti$"),
        basename(quanti_files))
]

length(demog_files)
length(quanti_files)

```

------------------------------------------------------------------------

# Demography: replicate cloud + mean

## Read and cache demography (stand totals)

```{r}
cfg_tag <- paste(CFGs, collapse = "-")
p_tag   <- paste(PCTs, collapse = "-")

demog_rds <- file.path(
  cache_dir,
  paste0("run_demog_long_cfg_", safe_slug(cfg_tag),
         "_p_", safe_slug(p_tag),
         "_", run_str,
         "_gmax", params$gen_max, ".rds")
)

cols_to_read <- c("replicate", "generation", "pop.tot", "a0.tot", "a1.tot", "a2.tot", "a3.tot")

if (isTRUE(params$use_cache) && file.exists(demog_rds)) {
  demog_long <- readRDS(demog_rds)
} else {
  dl <- vector("list", length(demog_files))

  for (i in seq_along(demog_files)) {
    f <- demog_files[i]
    meta <- parse_meta_from_basename(f)

    # filter early to save time
    if (!(meta$configuration %in% CFGs)) next
    if (!(meta$ori_proportion %in% PCTs)) next
    if (meta$layout_run != as.integer(params$layout_run)) next

    header <- names(fread(f, nrows = 0))
    keep <- intersect(cols_to_read, header)
    dt <- fread(f, select = keep)

    dt <- cbind(dt, meta)

    dl[[i]] <- melt(
      dt,
      id.vars = c("replicate","generation","configuration","ori_proportion","layout_run","k","b"),
      measure.vars = patterns("\\.tot$"),
      variable.name = "stage",
      value.name = "N"
    )
  }

  demog_long <- rbindlist(dl, use.names = TRUE, fill = TRUE)

  demog_long <- demog_long[!is.na(configuration)]
  demog_long <- demog_long[replicate %between% c(1, params$max_internal_replicate)]
  demog_long <- demog_long[generation <= params$gen_max]
  demog_long <- demog_long[configuration %in% CFGs & ori_proportion %in% PCTs]
  demog_long <- demog_long[layout_run == as.integer(params$layout_run)]

  if (isTRUE(params$use_cache)) saveRDS(demog_long, demog_rds)
}

stage_labs <- c(
  "pop.tot" = "Total population",
  "a0.tot"  = "Stage 0",
  "a1.tot"  = "Stage 1",
  "a2.tot"  = "Stage 2",
  "a3.tot"  = "Stage 3"
)
demog_long[, stage_label := stage_labs[stage]]
demog_long[, kb := paste0("k", k, " b", b)]

```

```{r}
demog_mean <- demog_long[
  , .(meanN = mean(N, na.rm = TRUE)),
  by = .(configuration, ori_proportion, generation, stage_label, k, b, kb)
]
```

## Plot: cloud (all stages)

```{r}
kb_levels <- sort(unique(demog_long$kb))

for (kb_i in kb_levels) {

  dL <- demog_long[kb == kb_i]
  dM <- demog_mean[kb == kb_i]

  # enforce facet order
  dL[, configuration := factor(configuration, levels = CFGs)]
  dM[, configuration := factor(configuration, levels = CFGs)]
  dL[, ori_proportion := factor(ori_proportion, levels = PCTs)]
  dM[, ori_proportion := factor(ori_proportion, levels = PCTs)]

  p_dem_cloud <- ggplot() +
    geom_line(
      data = dL,
      aes(x = generation, y = N,
          group = interaction(stage_label, replicate, configuration, ori_proportion),
          color = stage_label),
      linewidth = 0.15, alpha = 0.35
    ) +
    geom_line(
      data = dM,
      aes(x = generation, y = meanN, color = stage_label),
      linewidth = 0.8
    ) +
    facet_grid(ori_proportion ~ configuration, scales = "free_y") +
    labs(
      title = paste0("Demography cloud + mean (", kb_i, ")"),
      subtitle = paste0("Run ", run_str, " | configs: ", paste(CFGs, collapse=", "),
                        " | p: ", paste(PCTs, collapse=", "),
                        " | reps 1..", params$max_internal_replicate),
      x = "Generation", y = "Abundance", color = "Stage"
    ) +
    scale_y_continuous(labels = comma) +
    theme_bw()

  print(p_dem_cloud)
  save_plot(p_dem_cloud,
            paste0("run_demography_cloud_all_stages_", safe_slug(kb_i)),
            w = 18, h = 10)
}

```

## Plot: cloud (adults only)

```{r}
kb_levels <- sort(unique(demog_long$kb))

for (kb_i in kb_levels) {

  dL <- demog_long[kb == kb_i & stage_label == "Stage 3"]
  dM <- demog_mean[kb == kb_i & stage_label == "Stage 3"]

  dL[, configuration := factor(configuration, levels = CFGs)]
  dM[, configuration := factor(configuration, levels = CFGs)]
  dL[, ori_proportion := factor(ori_proportion, levels = PCTs)]
  dM[, ori_proportion := factor(ori_proportion, levels = PCTs)]

  p_dem_cloud_adults <- ggplot() +
    geom_line(
      data = dL,
      aes(x = generation, y = N,
          group = interaction(replicate, configuration, ori_proportion)),
      linewidth = 0.3, alpha = 0.35, color = "dodgerblue"
    ) +
    geom_line(
      data = dM,
      aes(x = generation, y = meanN),
      linewidth = 1.2, color = "dodgerblue"
    ) +
    facet_grid(ori_proportion ~ configuration, scales = "free_y") +
    labs(
      title = paste0("Adults only (Stage 3): cloud + mean (", kb_i, ")"),
      subtitle = paste0("Run ", run_str, " | reps 1..", params$max_internal_replicate),
      x = "Generation", y = "Adults (Stage 3)"
    ) +
    scale_y_continuous(labels = comma) +
    theme_bw() +
    theme(
      panel.background = element_rect(fill = "grey98", colour = NA),
      panel.grid.major = element_line(colour = "grey88", linewidth = 0.3),
      panel.grid.minor = element_line(colour = "grey92", linewidth = 0.2)
    )

  print(p_dem_cloud_adults)
  save_plot(p_dem_cloud_adults,
            paste0("run_demography_cloud_adults_only_", safe_slug(kb_i)),
            w = 18, h = 10)
}
```

------------------------------------------------------------------------

# Quanti (P1): distribution over time and hybrids

```{r}
p1_bins_rds    <- file.path(cache_dir, paste0("run_quanti_p1_bins_cfg_", safe_slug(cfg_tag),
                                              "_p_", safe_slug(p_tag),
                                              "_", run_str, "_gmax", params$gen_max, ".rds"))
p1_metrics_rds <- file.path(cache_dir, paste0("run_quanti_p1_metrics_cfg_", safe_slug(cfg_tag),
                                              "_p_", safe_slug(p_tag),
                                              "_", run_str, "_gmax", params$gen_max, ".rds"))

bw  <- as.numeric(params$p1_bin_width)
breaks <- seq(-1, 1, by = bw)
if (tail(breaks, 1) < 1) breaks <- c(breaks, 1)

tol <- as.numeric(params$p1_tol)
STAGE <- as.integer(params$quanti_stage)

build_quanti_summaries <- function() {
  bin_list <- list()
  met_list <- list()

  for (f in quanti_files) {
    meta <- parse_meta_from_basename(f)
    suf  <- parse_quanti_suffix(f)

    if (!(meta$configuration %in% CFGs)) next
    if (!(meta$ori_proportion %in% PCTs)) next
    if (meta$layout_run != as.integer(params$layout_run)) next
    if (suf$generation > params$gen_max) next

    dt <- fread(f, select = c("stage", "P1"))
    dt <- dt[stage == STAGE]
    if (nrow(dt) == 0) next

    dt[, P1 := as.numeric(P1)]

    dt[, is_f1    := abs(P1) <= tol]
    dt[, is_intro := (P1 > (-1 + tol)) & (P1 < (1 - tol))]

    N <- nrow(dt)

    met_list[[length(met_list) + 1]] <- cbind(
      data.table(
        generation = suf$generation,
        replicate  = suf$replicate,
        N = N,
        prop_f1 = mean(dt$is_f1, na.rm = TRUE),
        prop_intro = mean(dt$is_intro, na.rm = TRUE),
        mean_P1 = mean(dt$P1, na.rm = TRUE)
      ),
      meta
    )

    dt[, bin_id := findInterval(P1, breaks, rightmost.closed = TRUE, all.inside = TRUE)]
    bin_mids <- (breaks[-length(breaks)] + breaks[-1]) / 2
    dt[, bin_mid := bin_mids[bin_id]]

    bins <- dt[, .N, by = .(bin_mid)]
    bins[, prop := N / sum(N)]
    bins[, `:=`(generation = suf$generation, replicate = suf$replicate)]
    bins <- cbind(bins, meta)

    bin_list[[length(bin_list) + 1]] <- bins
  }

  list(
    p1_bins = rbindlist(bin_list, use.names = TRUE, fill = TRUE),
    p1_metrics = rbindlist(met_list, use.names = TRUE, fill = TRUE)
  )
}

if (isTRUE(params$use_cache) && file.exists(p1_bins_rds) && file.exists(p1_metrics_rds)) {
  p1_bins <- readRDS(p1_bins_rds)
  p1_metrics <- readRDS(p1_metrics_rds)
} else {
  tmp <- build_quanti_summaries()
  p1_bins <- tmp$p1_bins
  p1_metrics <- tmp$p1_metrics

  if (isTRUE(params$use_cache)) {
    saveRDS(p1_bins, p1_bins_rds)
    saveRDS(p1_metrics, p1_metrics_rds)
  }
}

p1_bins    <- p1_bins[replicate %between% c(1, params$max_internal_replicate)]
p1_metrics <- p1_metrics[replicate %between% c(1, params$max_internal_replicate)]

p1_bins    <- p1_bins[configuration %in% CFGs & ori_proportion %in% PCTs & layout_run == as.integer(params$layout_run)]
p1_metrics <- p1_metrics[configuration %in% CFGs & ori_proportion %in% PCTs & layout_run == as.integer(params$layout_run)]

p1_bins[, kb := paste0("k", k, " b", b)]
p1_metrics[, kb := paste0("k", k, " b", b)]

```

## Plot: P1 distribution over time (tile plot; adults)

Each tile shows the **mean proportion** (across internal replicates) of individuals whose P1 falls into that bin at that generation.

```{r}
bin_mids_expected <- (breaks[-length(breaks)] + breaks[-1]) / 2
bin_mids_expected <- round(bin_mids_expected, 2)
p1_bins[, bin_mid := round(as.numeric(bin_mid), 2)]

kb_levels_q <- sort(unique(p1_bins$kb))

for (kb_i in kb_levels_q) {

  pb <- p1_bins[kb == kb_i]
  if (nrow(pb) == 0) next

  # full grid so missing bins become 0
  keys <- unique(pb[, .(configuration, ori_proportion, layout_run, k, b, kb, generation, replicate)])

  grid <- keys[
    , .(bin_mid = bin_mids_expected),
    by = .(configuration, ori_proportion, layout_run, k, b, kb, generation, replicate)
  ]

  p1_bins_full <- merge(
    grid, pb,
    by = c("configuration","ori_proportion","layout_run","k","b","kb","generation","replicate","bin_mid"),
    all.x = TRUE
  )
  p1_bins_full[is.na(prop), prop := 0]

  p1_bin_mean <- p1_bins_full[
    , .(mean_prop = mean(prop)),
    by = .(configuration, ori_proportion, generation, bin_mid, k, b, kb)
  ]

  p1_bin_mean[, configuration := factor(configuration, levels = CFGs)]
  p1_bin_mean[, ori_proportion := factor(ori_proportion, levels = PCTs)]
  p1_bin_mean[, gen_f := factor(generation, levels = sort(unique(generation)))]

  p_p1_tile <- ggplot(p1_bin_mean, aes(x = gen_f, y = bin_mid, fill = mean_prop)) +
    geom_tile() +
    facet_grid(ori_proportion ~ configuration) +
    scale_fill_viridis_c(option = "C", trans = "sqrt") +
    labs(
      title = paste0("P1 distribution over time (stage ", STAGE, "; mean across internal reps) — ", kb_i),
      subtitle = paste0("Run ", run_str, " | p: ", paste(PCTs, collapse=", "), " | configs: ", paste(CFGs, collapse=", ")),
      x = "Generation",
      y = "P1 bin midpoint (-1 = sylvatica, 0 = hybrid, +1 = orientalis)",
      fill = "Mean\nproportion"
    ) +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      panel.background = element_rect(fill = "white", colour = NA)
    )

  print(p_p1_tile)
  save_plot(p_p1_tile,
            paste0("run_quanti_P1_distribution_tile_", safe_slug(kb_i)),
            w = 18, h = 10)
}


```

## Plot: How many hybrids over time (F1 = P1=0)

```{r}
kb_levels_m <- sort(unique(p1_metrics$kb))

for (kb_i in kb_levels_m) {

  pm <- p1_metrics[kb == kb_i]
  if (nrow(pm) == 0) next

  p1_f1_time <- pm[
    , .(mean_f1 = mean(prop_f1, na.rm = TRUE)),
    by = .(configuration, ori_proportion, generation, k, b, kb)
  ]

  p1_f1_time[, configuration := factor(configuration, levels = CFGs)]
  p1_f1_time[, ori_proportion := factor(ori_proportion, levels = PCTs)]

  p_f1 <- ggplot(p1_f1_time, aes(x = generation, y = mean_f1)) +
    geom_line(linewidth = 0.8) +
    facet_grid(ori_proportion ~ configuration) +
    labs(
      title = paste0("F1 hybrid proportion over time (P1=0) — ", kb_i),
      subtitle = paste0("Run ", run_str, " | mean across internal reps"),
      x = "Generation",
      y = "Proportion F1 (P1=0)"
    ) +
    theme_bw()

  print(p_f1)
  save_plot(p_f1,
            paste0("run_quanti_F1_prop_time_series_", safe_slug(kb_i)),
            w = 18, h = 10)
}

```
