---
title: "Simulation design and initialization"
subtitle: "NEMO-age hybridization scenarios on empty land"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
execute:
  warning: false
  message: true
---

This document generates all inputs required to run **(1)** a *burn-in* that produces a stable sylvatica baseline forest and **(2)** scenario-specific *planting simulations* initialized from that burn-in output by clearing selected patches and inserting orientalis seedlings. It writes the grid, dispersal matrices, initialization matrices, and INI files, and produces sanity plots of the planting layouts.

## Packages

```{r}
#| label: packages
#| echo: true

suppressPackageStartupMessages({
  library(terra)
  library(data.table)
  library(ggplot2)
  library(cowplot)
})
```

## NEMO setup

Create a shortcut to the location of the compiled NEMO-AGE binary.\
This facilitates calling NEMO in the current project root.

``` bash
ln -sf /home/jannis/nemoage/nemoage0.32.6b/bin/nemoageE.E.E ./nemoage
chmod +x ./nemoage
```

## Control panel

```{r}
#| label: control
#| echo: true

# Template (must exist)
TEMPLATE_INI <- "ini_files/templates/nemoage_template.ini"
stopifnot(file.exists(TEMPLATE_INI))

# Output folders
OUT <- list(
  input   = "input_files",
  ini     = "ini_files",
  results = "results",
  scripts = "scripts"
)

DIRS <- list(
  grid     = file.path(OUT$input, "grid"),
  cfg      = file.path(OUT$input, "configs_txt"),
  quanti   = file.path(OUT$input, "quanti_init_freq"),
  stage    = file.path(OUT$input, "patch_init_stage_size"),
  disperse = file.path(OUT$input, "disperse"),
  plots    = file.path(OUT$input, "plots"),
  ini_burnin = file.path(OUT$ini, "burnin"),
  ini_run    = file.path(OUT$ini, "run")
)

for (d in c(OUT, DIRS)) dir.create(d, showWarnings = FALSE, recursive = TRUE)

# Landscape
GRID_RES <- 4      # meters (patch = 4x4 m)
EXTENT_M <- 100    # 100x100 m
N        <- EXTENT_M / GRID_RES  # 25
CRS      <- "EPSG:3035"
GRID_SHP <- file.path(DIRS$grid, "Grid_4x4m_100x100m.shp")

# Scenario design
PROP_LEVELS    <- c(0.10, 0.25, 0.40)
CONFIG_LEVELS  <- c("dispersed", "one_cluster", "multi_cluster", "transects")
N_LAYOUT_RUNS  <- 5  # "biological replicates" per (config x proportion)

# Planting scheme for orientalis
PLANTING_SCHEME <- c(0, 20, 0, 0)  # stage0..stage3

# Burn-in + run settings
BURNIN_GENS <- 500
RUN_GENS    <- 500
K_VALUE     <- 30
B_VALUE     <- 0.03

# write stats + quanti every 50 generations
LOG_EVERY <- 50
LOG_TIMES <- c(2, seq(LOG_EVERY, RUN_GENS, by = LOG_EVERY))  # include generation 2

# NEMO replicates
NEMO_REPS_BURNIN <- 1
NEMO_REPS_RUN    <- 30

# Dispersal (fixed)
# SEED
SEED_D_MEAN <- 5
SEED_B      <- 1.3
SEED_D_T    <- 50
# POLLEN
POLLEN_D_MEAN <- 30
POLLEN_B      <- 1
POLLEN_D_T    <- 100

# Layout controls
# Dispersed (Poisson-disc-like): larger -> more regular spacing
DISPERSED_STRENGTH <- 1.20

# One cluster: how "soft" the square edge is (0 = perfect square)
CLUSTER_JITTER <- 0.15

# Multi clusters: number of clusters (3 or 4 are typical)
MULTI_N_CLUSTERS <- 4

# Transects: approximate number of 1-patch-wide transects for min/max proportions
TRANSECTS_MIN <- 2
TRANSECTS_MAX <- 7

# Seed (deterministic layout generation)
SEED_BASE <- 9999
```

## Grid

This creates the grid if it doesn't exist yet and computes patch centroids and the centroid distance matrix (meters).

```{r}
#| label: grid
#| echo: true

make_or_load_grid_strict <- function(shp_path, extent_m = 100, res = 4, crs = "EPSG:3035") {
  if (file.exists(shp_path)) return(vect(shp_path))

  r <- rast(extent = ext(0, extent_m, 0, extent_m), resolution = res, crs = crs)
  r[] <- 1:ncell(r)  # row-wise patchID in raster order

  grid_poly <- as.polygons(r, dissolve = FALSE)
  names(grid_poly)[names(grid_poly) == "lyr.1"] <- "patchID"
  writeVector(grid_poly, shp_path, filetype = "ESRI Shapefile", overwrite = TRUE)

  grid_poly
}

grid <- make_or_load_grid_strict(GRID_SHP, EXTENT_M, GRID_RES, CRS)
stopifnot(nrow(grid) == N * N)

cent <- centroids(grid)

coords <- as.data.frame(crds(cent))
names(coords) <- c("x", "y")
coords$patchID <- grid$patchID

# matrix-like plotting coords
coords$row <- ((coords$patchID - 1) %/% N) + 1
coords$col <- ((coords$patchID - 1) %%  N) + 1
coords$y_plot <- N - coords$row + 1
coords$x_plot <- coords$col

patch_ids <- coords$patchID
n_patches <- length(patch_ids)

dist_mat <- as.matrix(distance(cent))  # meters
```

## NEMO writers

```{r}
#| label: nemo-writers
#| echo: true

# general NEMO matrix writer: {{row1}\n{row2}\n...}
write.matrix.nemo <- function(mat, outfile) {
  cat("{", file = outfile)
  for (i in seq_len(nrow(mat))) {
    cat("{", file = outfile, append = TRUE)
    cat(mat[i, ], sep = ",", file = outfile, append = TRUE)
    cat("}\n", file = outfile, append = TRUE)
  }
  cat("}\n", file = outfile, append = TRUE)
}

# patch_nbfem vector file: {{K1,K2,...,Kn}}
write_patchK_vector <- function(K_vec, outfile) {
  stopifnot(length(K_vec) > 0)
  cat("{{", file = outfile)
  cat(K_vec, sep = ",", file = outfile, append = TRUE)
  cat("}}\n", file = outfile, append = TRUE)
}
```

## Dispersal matrices

```{r}
#| label: dispersal
#| echo: true

exp_power_kernel <- function(d, b, x) {
  a <- d * exp(lgamma(2 / b) - lgamma(3 / b))
  b * exp(-(x^b / a^b)) / (2 * pi * a^2 * exp(lgamma(2 / b)))
}

build.reduced.dispersal.matrices <- function(num_patch, distance_matrix, d, b, d_thresh) {
  stopifnot(nrow(distance_matrix) == num_patch, ncol(distance_matrix) == num_patch)

  rate_full <- matrix(0, nrow = num_patch, ncol = num_patch)
  conn      <- matrix(NA_integer_, nrow = num_patch, ncol = num_patch)
  rate_red  <- matrix(NA_real_,    nrow = num_patch, ncol = num_patch)

  for (i in 1:num_patch) {
    rate_full[i, ] <- exp_power_kernel(d, b, distance_matrix[i, ])
    ord <- order(rate_full[i, ], decreasing = TRUE)

    conn[i, ]     <- ord
    rate_red[i, ] <- rate_full[i, ord]

    to_remove <- which(rate_red[i, ] < d_thresh)
    if (length(to_remove)) {
      rate_red[i, to_remove] <- NA
      conn[i, to_remove]     <- NA
    }

    s <- sum(rate_red[i, ], na.rm = TRUE)
    if (s > 0) rate_red[i, !is.na(rate_red[i, ])] <- rate_red[i, !is.na(rate_red[i, ])] / s
  }

  list(dispersal_matrix = rate_full, connectivity_matrix = conn, rate_matrix = rate_red)
}

# Ensure no empty rows: connectivity starts with focal patch; rate starts with 1 if empty.
write.red.dispersal.matrix <- function(connectivity_matrix, rate_matrix, conn_file, rate_file) {
  # connectivity
  cat("{", file = conn_file)
  for (i in 1:nrow(connectivity_matrix)) {
    cat("{", file = conn_file, append = TRUE)
    cat(i, file = conn_file, append = TRUE)
    if (!all(is.na(connectivity_matrix[i, ]))) {
      non_na <- connectivity_matrix[i, which(!is.na(connectivity_matrix[i, ]))]
      non_na <- non_na[non_na != i]
      if (length(non_na)) {
        cat(",", file = conn_file, append = TRUE)
        cat(non_na, sep = ",", file = conn_file, append = TRUE)
      }
    }
    cat("}\n", file = conn_file, append = TRUE)
  }
  cat("}\n", file = conn_file, append = TRUE)

  # rate
  cat("{", file = rate_file)
  for (i in 1:nrow(rate_matrix)) {
    cat("{", file = rate_file, append = TRUE)
    if (all(is.na(rate_matrix[i, ]))) {
      cat(1, file = rate_file, append = TRUE)
    } else {
      cat(rate_matrix[i, which(!is.na(rate_matrix[i, ]))], sep = ",", file = rate_file, append = TRUE)
    }
    cat("}\n", file = rate_file, append = TRUE)
  }
  cat("}\n", file = rate_file, append = TRUE)
}

# thresholds (rate at distance d_t)
seed_thresh   <- exp_power_kernel(SEED_D_MEAN,   SEED_B,   SEED_D_T)
pollen_thresh <- exp_power_kernel(POLLEN_D_MEAN, POLLEN_B, POLLEN_D_T)

seed_mats <- build.reduced.dispersal.matrices(n_patches, dist_mat, SEED_D_MEAN, SEED_B, seed_thresh)
pol_mats  <- build.reduced.dispersal.matrices(n_patches, dist_mat, POLLEN_D_MEAN, POLLEN_B, pollen_thresh)

seed_conn_file   <- file.path(DIRS$disperse, sprintf("Seed_connectivity_matrix_d%s.txt", SEED_D_MEAN))
seed_rate_file   <- file.path(DIRS$disperse, sprintf("Seed_rate_matrix_d%s.txt", SEED_D_MEAN))
pollen_conn_file <- file.path(DIRS$disperse, sprintf("Pollen_connectivity_matrix_d%s.txt", POLLEN_D_MEAN))
pollen_rate_file <- file.path(DIRS$disperse, sprintf("Pollen_rate_matrix_d%s.txt", POLLEN_D_MEAN))

write.red.dispersal.matrix(seed_mats$connectivity_matrix, seed_mats$rate_matrix, seed_conn_file, seed_rate_file)
write.red.dispersal.matrix(pol_mats$connectivity_matrix,  pol_mats$rate_matrix,  pollen_conn_file, pollen_rate_file)

message("Wrote dispersal matrices to: ", DIRS$disperse)
```

## Burn-in inputs

```{r}
#| label: burnin-inputs
#| echo: true

# sylvatica-only quanti init: all zeros
q_burnin <- matrix(0, nrow = n_patches, ncol = 1)
q_burnin_file <- file.path(DIRS$quanti, "quanti_init_file_burnin_sylv.txt")
write.matrix.nemo(q_burnin, q_burnin_file)

# burn-in stage init: a small fraction of patches with one adult, plus uniform seedlings
build_patch_init_stage_size_burnin <- function(n_patches, adult_occupancy = 0.20, stage1_seedlings = 5, seed = 1) {
  set.seed(seed)
  mat <- matrix(0, nrow = n_patches, ncol = 4)
  colnames(mat) <- c("stage0","stage1","stage2","stage3")

  adults <- rep(0, n_patches)
  idx <- sample(seq_len(n_patches), size = as.integer(round(adult_occupancy * n_patches)))
  adults[idx] <- 1

  mat[, "stage3"] <- adults
  mat[, "stage1"] <- stage1_seedlings
  mat
}

burnin_stage <- build_patch_init_stage_size_burnin(n_patches, adult_occupancy = 0.20, stage1_seedlings = 5, seed = SEED_BASE)
burnin_stage_file <- file.path(DIRS$stage, "patch_init_stage_size_burnin_sylv.txt")
write.matrix.nemo(burnin_stage, burnin_stage_file)

message("Burn-in init files written:\n- ", q_burnin_file, "\n- ", burnin_stage_file)
```

## INI templating helpers

```{r}
#| label: ini-helpers
#| echo: true

read_ini_lines <- function(path) readLines(path, warn = FALSE)

set_ini_param <- function(lines, key, value) {
  rx <- paste0("^\\s*", key, "\\b")
  idx <- grep(rx, lines)

  newline <- paste(key, value)

  # key not present -> append
  if (length(idx) == 0) return(c(lines, newline))

  # replace first occurrence
  lines[idx[1]] <- newline

  # remove any additional occurrences
  if (length(idx) > 1) lines <- lines[-idx[-1]]

  lines
}

# drop lines matching regex
drop_ini_lines <- function(lines, regex) lines[!grepl(regex, lines)]
```

## Write the burn-in INI

```{r}
#| label: write-burnin-ini
#| echo: true

burnin_lines <- read_ini_lines(TEMPLATE_INI)

# core
burnin_lines <- set_ini_param(burnin_lines, "replicates", NEMO_REPS_BURNIN)
burnin_lines <- set_ini_param(burnin_lines, "generations", BURNIN_GENS)

# outputs
burnin_root <- file.path(OUT$results, "burnin")
dir.create(burnin_root, showWarnings = FALSE, recursive = TRUE)
burnin_name <- sprintf("burnin_k%d_b%.3g", K_VALUE, B_VALUE)
burnin_lines <- set_ini_param(burnin_lines, "root_dir", burnin_root)
burnin_lines <- set_ini_param(burnin_lines, "filename", burnin_name)
burnin_lines <- set_ini_param(burnin_lines, "logfile", paste0(burnin_name, "_log"))

# init
burnin_lines <- set_ini_param(burnin_lines, "quanti_init_freq", paste0("&", q_burnin_file))
burnin_lines <- set_ini_param(burnin_lines, "patch_init_stage_size", paste0("&", burnin_stage_file))

# K and competition (fixed)
burnin_lines <- set_ini_param(burnin_lines, "patch_nbfem", K_VALUE)
burnin_lines <- set_ini_param(burnin_lines, "regulation_by_competition", B_VALUE)

# logging
burnin_lines <- set_ini_param(burnin_lines, "stat_log_time",
                              paste0("{{", paste(LOG_TIMES, collapse = ","), "}}"))
# quanti output in burn-in too
burnin_lines <- set_ini_param(burnin_lines, "quanti_dir", "quanti")
burnin_lines <- set_ini_param(burnin_lines, "quanti_logtime",
                              paste0("{{", paste(LOG_TIMES, collapse = ","), "}}"))

# dispersal
burnin_lines <- set_ini_param(burnin_lines, "seed_disperse_connectivity_matrix", paste0("&", seed_conn_file))
burnin_lines <- set_ini_param(burnin_lines, "seed_disperse_reduced_matrix",       paste0("&", seed_rate_file))
burnin_lines <- set_ini_param(burnin_lines, "breed_disperse_connectivity_matrix_mal", paste0("&", pollen_conn_file))
burnin_lines <- set_ini_param(burnin_lines, "breed_disperse_reduced_matrix_mal",       paste0("&", pollen_rate_file))

burnin_ini <- file.path(DIRS$ini_burnin, paste0(burnin_name, ".ini"))
writeLines(burnin_lines, burnin_ini)

# burnin output txt we will parse later
BURNIN_TXT <- file.path(burnin_root, paste0(burnin_name, ".txt"))

message("Wrote burn-in ini: ", burnin_ini)
message("Burn-in will write: ", BURNIN_TXT)
```

### Run the burn-in

Run this once from the project root:

``` bash
./nemoage ini_files/burnin/burnin_k30_b0.03.ini
```

## Build the baseline stage matrix from the burn-in output

We read the per-patch female counts (`a0.fem.p1 ... a3.fem.p625`) at the final generation and convert them into a `patch_init_stage_size` matrix for the scenario simulations.

```{r}
#| label: baseline-from-burnin
#| echo: true

extract_baseline_stage_matrix <- function(txt_path, n_patches, gen, rep = 1) {
  stopifnot(file.exists(txt_path))

  hdr <- names(fread(txt_path, nrows = 0))
  hdr2 <- gsub("^off\\.", "a0.", hdr) # older outputs may use off.* for stage0

  stage_cols <- lapply(0:3, function(s) grep(paste0("^a", s, "\\.fem\\.p\\d+$"), hdr2, value = TRUE))
  if (any(lengths(stage_cols) == 0)) stop("Could not find per-patch female columns a0..a3 in: ", txt_path)

  cols_needed <- c("replicate", "generation", unlist(stage_cols))
  dt <- fread(txt_path, select = cols_needed, showProgress = FALSE)
  setnames(dt, gsub("^off\\.", "a0.", names(dt)))

  row <- dt[replicate == rep & generation == gen]
  if (nrow(row) != 1) stop("Could not find exactly one row for replicate=", rep, " generation=", gen)

  out <- matrix(0, nrow = n_patches, ncol = 4)
  colnames(out) <- paste0("stage", 0:3)

  for (s in 0:3) {
    cols <- grep(paste0("^a", s, "\\.fem\\.p\\d+$"), names(row), value = TRUE)
    pnum <- as.integer(sub(paste0("^a", s, "\\.fem\\.p"), "", cols))
    ord  <- order(pnum)
    vals <- as.numeric(unlist(row[, ..cols])[ord])
    out[pnum[ord], s + 1] <- vals
  }

  out
}

BASELINE_GEN <- BURNIN_GENS
BASELINE_REP <- 1

if (!file.exists(BURNIN_TXT)) {
  stop("Burn-in output not found yet: ", BURNIN_TXT, "\nRun burn-in first, then re-render or re-run this section.")
}

baseline_stage <- extract_baseline_stage_matrix(BURNIN_TXT, n_patches, gen = BASELINE_GEN, rep = BASELINE_REP)
baseline_stage_file <- file.path(DIRS$stage, sprintf("patch_init_stage_size_baseline.txt"))
write.matrix.nemo(baseline_stage, baseline_stage_file)

message("Wrote baseline stage init: ", baseline_stage_file)
```

## Planting layout generators

All functions return a vector of `patchID` to be planted as orientalis.

```{r}
#| label: layout-generators
#| echo: true

# utility: square layer distance (Chebyshev) around a center in (row,col)
rank_by_square_layers <- function(rows, cols, center_r, center_c, jitter = 0, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  d <- pmax(abs(rows - center_r), abs(cols - center_c))
  if (jitter > 0) d <- d + runif(length(d), -jitter, jitter)

  order(d)
}

# 1) dispersed: Poisson-disc-like sequential selection with min-distance relaxation
select_orientalis_dispersed <- function(
  nO,
  patch_ids,
  dist_mat,
  base_spacing_m,
  seed = NULL,
  strength = 1.20,
  relax_step = 0.95,
  max_relax = 12
) {
  if (!is.null(seed)) set.seed(seed)
  if (nO <= 0) return(integer(0))
  if (nO >= length(patch_ids)) return(patch_ids)

  minDist <- base_spacing_m * strength

  for (attempt in seq_len(max_relax)) {
    chosen <- integer(0)

    for (cand in sample(patch_ids)) {
      if (!length(chosen) || min(dist_mat[cand, chosen]) >= minDist) {
        chosen <- c(chosen, cand)
      }

      if (length(chosen) >= nO) return(chosen[1:nO])
    }

    minDist <- minDist * relax_step
  }

  # fallback
  sample(patch_ids, nO)
}

# 2) one square-ish cluster around (near) center, with slight edge softness via jitter
select_orientalis_one_cluster <- function(nO, coords_df, seed = NULL, jitter = 0.15) {
  if (!is.null(seed)) set.seed(seed)
  if (nO <= 0) return(integer(0))
  if (nO >= nrow(coords_df)) return(coords_df$patchID)

  # random small shift around center to create different layout runs
  center_r <- round(N / 2) + sample(-2:2, 1)
  center_c <- round(N / 2) + sample(-2:2, 1)

  center_r <- max(1, min(N, center_r))
  center_c <- max(1, min(N, center_c))

  ord <- rank_by_square_layers(
    coords_df$row,
    coords_df$col,
    center_r,
    center_c,
    jitter = jitter,
    seed = seed
  )

  coords_df$patchID[ord][1:nO]
}

# 3) multiple square-ish clusters with roughly equal size, placed roughly evenly across the grid
select_orientalis_multi_cluster <- function(
  nO,
  coords_df,
  seed = NULL,
  n_centers = 4,
  jitter = 0.15
) {
  if (!is.null(seed)) set.seed(seed)
  if (nO <= 0) return(integer(0))
  if (nO >= nrow(coords_df)) return(coords_df$patchID)

  n_centers <- max(2, min(n_centers, 6))

  # choose center locations by partitioning the grid into n_centers regions
  if (n_centers == 3) {
    centers_rc <- rbind(
      c(round(N * 0.25), round(N * 0.25)),
      c(round(N * 0.75), round(N * 0.25)),
      c(round(N * 0.50), round(N * 0.75))
    )
  } else { # default 4
    centers_rc <- rbind(
      c(round(N * 0.25), round(N * 0.25)),
      c(round(N * 0.75), round(N * 0.25)),
      c(round(N * 0.25), round(N * 0.75)),
      c(round(N * 0.75), round(N * 0.75))
    )

    if (n_centers > 4) {
      extra <- replicate(
        n_centers - 4,
        c(sample(5:(N - 4), 1), sample(5:(N - 4), 1))
      )
      centers_rc <- rbind(centers_rc, t(extra))
    }
  }

  # jitter centers slightly for each run
  centers_rc <- centers_rc +
    matrix(sample(-2:2, nrow(centers_rc) * 2, replace = TRUE), ncol = 2)

  centers_rc[, 1] <- pmin(N, pmax(1, centers_rc[, 1]))
  centers_rc[, 2] <- pmin(N, pmax(1, centers_rc[, 2]))

  # sizes per cluster (near-equal)
  base  <- floor(nO / nrow(centers_rc))
  sizes <- rep(base, nrow(centers_rc))
  sizes[1:(nO - sum(sizes))] <- sizes[1:(nO - sum(sizes))] + 1

  chosen    <- integer(0)
  available <- coords_df$patchID

  for (k in seq_len(nrow(centers_rc))) {
    center_r <- centers_rc[k, 1]
    center_c <- centers_rc[k, 2]

    sub <- coords_df[match(available, coords_df$patchID), ]

    ord <- rank_by_square_layers(
      sub$row,
      sub$col,
      center_r,
      center_c,
      jitter = jitter,
      seed = seed + k
    )

    take <- sub$patchID[ord][1:min(sizes[k], length(ord))]

    chosen    <- c(chosen, take)
    available <- setdiff(available, take)

    if (length(chosen) >= nO) break
  }

  chosen <- unique(chosen)

  if (length(chosen) > nO) {
    chosen <- chosen[1:nO]
  }

  if (length(chosen) < nO) {
    chosen <- c(
      chosen,
      sample(setdiff(coords_df$patchID, chosen), nO - length(chosen))
    )
  }

  chosen
}

# 4) transects (hard edges): full-length straight columns, approx matching p
select_orientalis_transects <- function(
  nO, coords_df, prop_orientalis = NULL, seed = NULL,
  min_tr = NULL, max_tr = NULL, gap = 4L, ...
) {
  if (!is.null(seed)) set.seed(seed)
  if (nO <= 0) return(integer(0))
  if (nO >= nrow(coords_df)) return(coords_df$patchID)

  N <- max(coords_df$row)   # local grid size
  k <- 3L

  n_full <- nO %/% N
  rem    <- nO %% N

  # distribute FULL columns across 3 transects
  f <- rep(n_full %/% k, k)
  r <- n_full %% k
  if (r > 0) f[sample.int(k, r)] <- f[sample.int(k, r)] + 1L

  # choose which transect gets the ONE partial column (if any)
  w_used <- f
  t_rem  <- 0L
  if (rem > 0) {
    z <- which(f == 0L)
    t_rem <- if (length(z)) sample(z, 1) else sample.int(k, 1)
    w_used[t_rem] <- w_used[t_rem] + 1L
  }

  w <- pmax(1L, w_used)  # keep 3 visible blocks (some may be unused at very low nO)

  slack <- N - (sum(w) + 2L * gap)
  if (slack < 0) stop("Impossible: N too small for 3 transects with this gap and nO.")

  pad <- as.vector(rmultinom(1, slack, rep(1, 4)))
  s1 <- 1L + pad[1]
  s2 <- s1 + w[1] + gap + pad[2]
  s3 <- s2 + w[2] + gap + pad[3]
  starts <- c(s1, s2, s3)

  chosen <- integer(0)

  for (t in 1:k) {
    cols_block <- starts[t]:(starts[t] + w[t] - 1L)
    cols_use   <- if (w_used[t] > 0) cols_block[seq_len(w_used[t])] else integer(0)

    if (f[t] > 0) {
      chosen <- c(chosen, coords_df$patchID[coords_df$col %in% cols_use[1:f[t]]])
    }

    if (rem > 0 && t == t_rem) {
      pc   <- cols_use[f[t] + 1L]
      rows <- if (sample(c(TRUE, FALSE), 1)) 1:rem else (N - rem + 1L):N
      chosen <- c(chosen, coords_df$patchID[coords_df$col == pc & coords_df$row %in% rows])
    }
  }

  chosen
}

# master wrapper
select_orientalis <- function(
  nO,
  configuration,
  patch_ids,
  dist_mat,
  coords_df,
  prop_orientalis,
  seed = NULL
) {
  configuration <- match.arg(configuration, CONFIG_LEVELS)

  if (configuration == "dispersed") {
    spacing_m <- sqrt(length(patch_ids) / max(1, nO)) * GRID_RES

    return(
      select_orientalis_dispersed(
        nO,
        patch_ids,
        dist_mat,
        base_spacing_m = spacing_m,
        seed = seed,
        strength = DISPERSED_STRENGTH
      )
    )
  }

  if (configuration == "one_cluster") {
    return(
      select_orientalis_one_cluster(
        nO,
        coords_df,
        seed = seed,
        jitter = CLUSTER_JITTER
      )
    )
  }

  if (configuration == "multi_cluster") {
    return(
      select_orientalis_multi_cluster(
        nO,
        coords_df,
        seed = seed,
        n_centers = MULTI_N_CLUSTERS,
        jitter = CLUSTER_JITTER
      )
    )
  }

  if (configuration == "transects") {
    return(
      select_orientalis_transects(
        nO,
        coords_df,
        prop_orientalis = prop_orientalis,
        seed = seed,
        min_tr = TRANSECTS_MIN,
        max_tr = TRANSECTS_MAX
      )
    )
  }

  stop("Unknown configuration: ", configuration)
}
```

## Scenario writers + plotting helpers

```{r}
#| label: scenario-writers
#| echo: true

scenario_id <- function(configuration, prop_orientalis, run_id) {
  sprintf("%s_p%02d_r%02d", configuration, round(100 * prop_orientalis), run_id)
}

build_config_table <- function(orientalis_ids, patch_ids) {
  dt <- data.table(patchID = patch_ids, patch_value = "S")
  dt[patchID %in% orientalis_ids, patch_value := "O"]
  dt[order(patchID)]
}

write_config_quarto_schema <- function(conf_dt, sim_name, outfile) {
  out <- copy(conf_dt)
  out[, simulation := sim_name]
  setcolorder(out, c("simulation", "patchID", "patch_value"))
  fwrite(out, outfile, sep = "\t")
}

# quanti init (one column): 1 in O patches, 0 otherwise
build_quanti_init <- function(conf_dt, n_patches) {
  mat <- matrix(0, nrow = n_patches, ncol = 1)
  mat[conf_dt$patchID, 1] <- ifelse(conf_dt$patch_value == "O", 1, 0)
  mat
}

# patch_init_stage_size for run: start from baseline, clear O patches, plant orientalis scheme
build_patch_init_stage_size_from_baseline <- function(conf_dt, baseline_stage_mat, scheme = c(0,20,0,0)) {
  stopifnot(nrow(baseline_stage_mat) == nrow(conf_dt))
  out <- baseline_stage_mat
  O <- conf_dt$patchID[conf_dt$patch_value == "O"]
  out[O, ] <- 0
  out[O, ] <- matrix(rep(scheme, each = length(O)), nrow = length(O), byrow = FALSE)
  out
}

plot_grid_binary_gg <- function(conf_dt, coords_df, title = NULL) {
  df <- merge(coords_df, conf_dt, by = "patchID", all.x = TRUE)
  df$patch_value <- ifelse(is.na(df$patch_value), "S", df$patch_value)

  ggplot(df, aes(x = x_plot, y = y_plot, fill = patch_value)) +
    geom_tile(color = "grey85", linewidth = 0.1) +
    coord_equal() +
    labs(title = title, x = "", y = "", fill = "") +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      legend.position = "none",
      plot.title = element_text(size = 10, hjust = 0.5))
}
```

## Generate scenarios (layouts + init matrices)

```{r}
#| label: generate-scenarios
#| echo: true

manifest <- data.table()

for (p in PROP_LEVELS) {
  nO_target <- p * n_patches
  nO <- as.integer(round(nO_target))
  p_used <- nO / n_patches

  for (cfg in CONFIG_LEVELS) {
    for (r in seq_len(N_LAYOUT_RUNS)) {

      sim_name <- scenario_id(cfg, p, r)

      seed <- SEED_BASE + round(p * 1000) * 1000 + match(cfg, CONFIG_LEVELS) * 100 + r

      O_ids <- select_orientalis(
        nO = nO,
        configuration = cfg,
        patch_ids = patch_ids,
        dist_mat = dist_mat,
        coords_df = coords,
        prop_orientalis = p,
        seed = seed
      )

      conf <- build_config_table(O_ids, patch_ids)

      # configs_txt
      cfg_file <- file.path(DIRS$cfg, paste0(sim_name, ".txt"))
      write_config_quarto_schema(conf, sim_name, cfg_file)

      # quanti init
      q_mat <- build_quanti_init(conf, n_patches)
      q_file <- file.path(DIRS$quanti, paste0("quanti_init_file_", sim_name, ".txt"))
      write.matrix.nemo(q_mat, q_file)

      # run stage init (baseline + clearing + planting)
      plant_mat <- build_patch_init_stage_size_from_baseline(conf, baseline_stage, scheme = PLANTING_SCHEME)
      stage_file <- file.path(DIRS$stage, paste0("patch_init_stage_size_", sim_name, ".txt"))
      write.matrix.nemo(plant_mat, stage_file)

      manifest <- rbind(
        manifest,
        data.table(
          sim_name = sim_name,
          configuration = cfg,
          prop_orientalis = p,
          prop_used = p_used,
          n_orientalis = nO,
          n_target = nO_target,
          run_id = r,
          seed = seed,
          cfg_file = cfg_file,
          quanti_file = q_file,
          stage_file = stage_file
        )
      )
    }
  }
}

manifest_file <- file.path(OUT$input, "manifest_scenarios.tsv")
fwrite(manifest, manifest_file, sep = "\t")
message("Generated ", nrow(manifest), " scenario layouts.")
message("Manifest: ", manifest_file)
```

## Layout plots

```{r}
#| label: scenario-plots
#| echo: true

PLOT_DIR <- file.path(DIRS$plots, "layout_runs")
dir.create(PLOT_DIR, showWarnings = FALSE, recursive = TRUE)

manifest_dt <- copy(manifest)
invisible(manifest_dt[, scenario_key := sub("_r[0-9]+$", "", sim_name)])

scenario_groups <- manifest_dt[
  ,
  .(
    cfg_files = list(cfg_file),
    sims = list(sim_name),
    prop = prop_orientalis[1],
    configuration = configuration[1],
    n_orientalis = n_orientalis[1]
  ),
  by = .(scenario_key)
]

for (i in seq_len(nrow(scenario_groups))) {
  key <- scenario_groups$scenario_key[i]
  cfg <- scenario_groups$configuration[i]
  p   <- scenario_groups$prop[i]

  cfg_files <- unlist(scenario_groups$cfg_files[i])
  sims      <- unlist(scenario_groups$sims[i])

  ord <- order(sims)
  cfg_files <- cfg_files[ord]
  sims <- sims[ord]
  
  plist <- lapply(seq_along(cfg_files), function(j) {
    dt <- fread(cfg_files[j])[, .(patchID, patch_value)]
    nO_run <- sum(dt$patch_value == "O")
    rep_lab <- sprintf("r%02d | %d (%.2f%%)", j, nO_run, 100 * nO_run / n_patches)
    plot_grid_binary_gg(dt, coords, rep_lab)
  })

  nO_used <- scenario_groups$n_orientalis[i]
  main_title <- sprintf(
    "%s | requested %.0f%% (%.1f/%d) | used %d/%d = %.2f%%",
    cfg,
    100 * p,
    p * n_patches,
  n_patches,
  nO_used,
  n_patches,
  100 * nO_used / n_patches
)


  panel <- cowplot::plot_grid(plotlist = plist, nrow = 1, align = "hv")
  final <- cowplot::plot_grid(
    cowplot::ggdraw() + cowplot::draw_label(main_title, fontface = "bold", size = 14),
    panel,
    ncol = 1,
    rel_heights = c(0.18, 1)
  )
  
  final <- final + theme(plot.background = element_rect(fill = "white", color = NA))

  out_png <- file.path(PLOT_DIR, paste0("overview_", key, ".png"))
  ggsave(out_png, final, width = 16, height = 3.8, dpi = 220)
}

message("Saved layout overviews in: ", PLOT_DIR)
```

## Write run INIs (scenario simulations)

Each scenario run starts from `baseline_stage` with scenario-specific cleared+planted patches and the scenario-specific quanti init file.

```{r}
#| label: write-run-inis
#| echo: true

run_template <- read_ini_lines(TEMPLATE_INI)

# set constants
run_template <- set_ini_param(run_template, "replicates", NEMO_REPS_RUN)
run_template <- set_ini_param(run_template, "generations", RUN_GENS)

run_template <- set_ini_param(run_template, "patch_nbfem", K_VALUE)
run_template <- set_ini_param(run_template, "regulation_by_competition", B_VALUE)

run_template <- set_ini_param(run_template, "stat_log_time",
                              paste0("{{", paste(LOG_TIMES, collapse = ","), "}}"))
run_template <- set_ini_param(run_template, "quanti_dir", "quanti")
run_template <- set_ini_param(run_template, "quanti_logtime",
                              paste0("{{", paste(LOG_TIMES, collapse = ","), "}}"))

# dispersal
run_template <- set_ini_param(run_template, "seed_disperse_connectivity_matrix", paste0("&", seed_conn_file))
run_template <- set_ini_param(run_template, "seed_disperse_reduced_matrix",       paste0("&", seed_rate_file))
run_template <- set_ini_param(run_template, "breed_disperse_connectivity_matrix_mal", paste0("&", pollen_conn_file))
run_template <- set_ini_param(run_template, "breed_disperse_reduced_matrix_mal",       paste0("&", pollen_rate_file))

ini_manifest <- data.table()

for (i in seq_len(nrow(manifest))) {
  sim  <- manifest$sim_name[i]
  cfg  <- manifest$configuration[i]
  p    <- manifest$prop_orientalis[i]
  rid  <- manifest$run_id[i]
  seed <- manifest$seed[i]

  # directory grouping
  run_root <- file.path(OUT$results, cfg, sprintf("p%02d", round(100*p)), sprintf("r%02d", rid), "run")
  dir.create(run_root, showWarnings = FALSE, recursive = TRUE)

  ini <- run_template
  ini <- set_ini_param(ini, "root_dir", run_root)
  ini <- set_ini_param(ini, "filename", paste0(sim, "_k", K_VALUE, "_b", B_VALUE))
  ini <- set_ini_param(ini, "logfile",  paste0(sim, "_k", K_VALUE, "_b", B_VALUE, "_log"))

  # init (baseline-derived stage matrix + scenario quanti init)
  ini <- set_ini_param(ini, "quanti_init_freq", paste0("&", manifest$quanti_file[i]))
  ini <- set_ini_param(ini, "patch_init_stage_size", paste0("&", manifest$stage_file[i]))

  ini_path <- file.path(DIRS$ini_run, paste0(sim, "_k", K_VALUE, "_b", B_VALUE, ".ini"))
  writeLines(ini, ini_path)

  ini_manifest <- rbind(
    ini_manifest,
    data.table(
      sim_name = sim,
      configuration = cfg,
      prop_orientalis = p,
      run_id = rid,
      ini = ini_path,
      run_root = run_root
    )
  )
}

ini_manifest_file <- file.path(OUT$scripts, "ini_manifest.tsv")
fwrite(ini_manifest, ini_manifest_file, sep = "\t")
message("Wrote run inis to: ", DIRS$ini_run)
message("INI manifest: ", ini_manifest_file)
```

## Write a run-all script

```{r}
#| label: write-run-all
#| echo: true

cmds <- sprintf("./nemoage %s", ini_manifest$ini)

sh_all <- file.path(OUT$scripts, "run_all.sh")
writeLines(c("#!/usr/bin/env bash", "set -euo pipefail", "", cmds), sh_all)
Sys.chmod(sh_all, "0755")

message("Wrote: ", sh_all)
```

Run all scenario simulations (after the burn-in step has been completed and baseline has been built):

``` bash
./scripts/run_all.sh
```

Move logs from root after running:

``` bash
./scripts/move_logs.sh
```
