---
title: "Hybridization_project"
author: "Camilla Stefanini"
format: html
editor: visual
execute:
    echo: false
    warning: false
    message: false
    freeze: false
editor_options: 
  chunk_output_type: console
---

## Introduction

We want to simulate different hypothetical scenarios of hybridization between European beech (*Fagus sylvatica)* and Oriental beech *(Fagus orientalis*) starting from different configurations. Oriental beech has already been introduced 100 years ago in Europe (see kurz et al, 2023), and today there is increasing attention to this species, which could improve genetic diversity to European beech. (more about E. beech and drought stress to add here).

These configurations come from a questionnaire which was given to foresters and researchers during the workshop about the introduction of Oriental beech species in Europe. They were asked to imagine that they own a land, and to indicate how they would plant (or cut) European beech or Oriental beech trees in 3 types of lands of 100 x 100 m each (each patch is 4x4 m), if their goal is to **maximise the hybridization between the two species**. In practice, they needed to draw the planting scheme they considered appropriate to achieve this goal, and indicate the planting method and the harvesting scheme.

**Goal of the project: reproduce what foresters and researchers indicated as optimal strategy, using simulations, and see what is the outcome.**

**Research question: which planting design maximizes hybridization and its persistence through time?**

Which planting scheme is giving more hybrids, and how this is influenced by the initial number of Orientalis introduced?

## Pipeline

1.  Find common patterns in the simulation scenarios and cluster the designs: translate each questionnaires into a small set of quantitative descriptors, for instance % orientalis (choose three values from the questionnaires), number of clusters (choose one or 2 values from the questionnaires), cluster size, level of aggregation, etc...

2.  Fix demographic and biological data based on literature ---\> (add a quick sensitivity analysis for the supplementary on carrying capacity and competition coefficient, by keeping only one value of % orientalis introduced)

3.  Create the three spatial configurations: **mixed**, **clustered** and **striped**? with 3 values of %orientalis introduced.

-\> in total, 3 x 3 = 9 spatial configurations? -\> NB: testing only set of simulations starting with European beech?

4.  For each spatial configuration, create the input files (matrices) and the init files.

5.  Run the simulations

-\> REPLICATES: **first exploratory test with 3-5 replicates, final simulations with 30 replicates, after everything is defined** -\> GENERATIONS (years): burnin simulations approx 20 to let E. beech grow and reproduce a bit, then actual simulations for 300-500 years.

6.  Analyse results

------------------------------------------------------------------------

Conversion from questionnaire to simulation:

| Questionnaire    |                                  NEMO |
|------------------|--------------------------------------:|
| 4x4m cell        |                                 patch |
| Tree presence    | patch carrying capacity (patch_nbfem) |
| Species identity |           quantitative trait (quanti) |
| Planting scheme  |                 patch_init_stage_size |
| Harvesting       |                                     ? |

The simulation should be defined at patch-level. In general: *patch_nbfem* = 1 for adults females = for a patch 4x4 there is only one adult.

## Simulation sets

### 1. Starting from empty land (NB: MAYBE NOT RELEVANT)

**Goal**\
Introduce *Fagus sylvatica* and *Fagus orientalis* seedlings into empty habitat in a single-step simulation.

**Simulation structure** - Single simulation (no burn-in) - Seedlings are planted directly according to carrying capacity and species frequency

**Key parameters**

-   **Patch availability**
    -   `patch_nbfem = X` everywhere\
        → seedlings can establish in all patches
-   **Initial stage structure**
    -   `patch_init_stage_size {{0, 20, 0, 0}}`\
        → \~20 seedlings planted per patch\
        → build corresponding matrix
-   **Species frequency**
    -   `quanti_init_freq = 0 or 1` depending on the species\
        → build matrix

------------------------------------------------------------------------

### 2. Starting from mature European beech forest (40 years)

This scenario uses **two simulations**: a burn-in to create the forest, followed by the actual simulation. Both can be run with the following bash code

``` bash
./nemoage ./ini_files/mixed1_burnin.ini ./ini_files/mixed1.ini
```

#### 2.1 Burn-in simulation: European beech forest

**Goal**\
Create a mature *F. sylvatica* forest and clear selected patches for later *F. orientalis* introduction.

**Duration**

-   20 years

**Key parameters**

-   **Initial stage structure**

    -   `patch_init_stage_size {{0, 0, 0, 1}}`\
        → single adult *sylvatica* per patch as in beech0 configuration\
        → beech0 matrix

-   **Species frequency**

    -   `quanti_init_freq = 0` everywhere

-   **Patch availability and cutting**

```         
-   `patch_nbfem`  (@g0 200, @g50  &patch_nbfem_cutting_sim_name_repX.txt)
        → seedlings can grow everywhere, but at the last generation of the simulation, the patches of the cutting are cleared  → **each replicate has to be linked to the right ini file** ???
```

-   **Output**

    -   Population stored using `STORE`\
        → see Nemoage manual (e.g. p. 106)

#### 2.2 Actual simulation: replacement with *Fagus orientalis*

**Goal**\
Introduce *F. orientalis* into selected patches and simulate replacement dynamics.

**Steps**

1.  **Load burn-in population**

    -   `source_pop = burnin binary file`

2.  **Plant *F. orientalis***

    -   `patch_init_stage_size {{0, X, 0, 1}}`\
        → planting patternlinked to the final patch_nbfem of the burnin for the cutting scheme
    -   `quanti_init_freq = 1` in planted patches\
        → planting pattern as defined in external matrices

3.  **Patch availability**

    -   `patch_nbfem` homogeneous\
        → no restriction on seedling establishment or dispersal

4.  **Post-establishment dynamics**

    -   Seedlings spread via dispersal\
        → fill available slots below `patch_nbfem`

    -   Adult mortality increased\
        → remove approximately 20% of adults\
        → exact value depends on the scenario?

------------------------------------------------------------------------

### 3. Starting from mixed mature forest (40 years) (NB: MAYBE WE DO NOT SIMULATE)

Same structure as Scenario 2, but with a mixed-species background forest.

``` bash
./nemoage ./ini_files/mixed1_burnin.ini ./ini_files/mixed1.ini
```

#### 3.1 Burn-in simulation: mixed forest

**Goal**\
Create a mature mixed forest and clear selected patches for *Fagus orientalis* introduction.

**Duration**

-   20 years

**Key parameters**

-   **Initial stage structure**
    -   `patch_init_stage_size {{0, 0, 0, 1}}`\
        → single adult *F. sylvatica* per patch\
        → beech0 matrix
-   **Species frequency**
    -   `quanti_init_freq = 0` everywhere
-   **Patch availability**
    -   `patch_nbfem {{X, 0, ...}}`
        -   `X` = available patches\
        -   `0` = *Abies* patches only\
            → build corresponding matrix

::: callout-note
**TODO**\
Clarify whether the competition coefficient **b** should be higher in mixed stands.
:::

-   **Output**
    -   Population stored using `STORE`

#### 3.2 Actual simulation: *Fagus orientalis* introduction

**Goal**\
Introduce *F. orientalis* into selected patches within a mixed-stand background and simulate replacement dynamics.

**Steps**

1.  **Load burn-in population**

    -   `source_pop = burnin binary file`

2.  **Plant *F. orientalis***

    -   `patch_init_stage_size {{0, X, 0, 1}}`
    -   `quanti_init_freq = 1` in planted patches\
        → planting pattern as defined in `config.txt`

3.  **Patch availability**

    -   `patch_nbfem {{X, 0, ...}}`\
        → *Abies* patches remain unavailable for beech establishment

4.  **Post-establishment dynamics**

    -   *Abies* continues occupying its patches
    -   Seedlings spread via dispersal\
        → fill available slots below `patch_nbfem`
    -   Adult mortality increased\
        → remove approximately 20% of adults\
        → exact value depends on the scenario

------------------------------------------------------------------------

## Starting values (demographic parameters)

Fixed (species-specific) parameters: (but potentially add in the Supp. a little sensitivity analysis on carryinc capacity and competition coeff. to check the robustness of the results)

1.  population structure and age of reproduction (40) as average value found in the literature (between 30 and 50), or 50 as found in https://www.cabidigitallibrary.org/doi/full/10.5555/19531101078

    pop_age_structure {{0,1,4,40}}

    Stage 0: offspring of year 1

    Stage 1: offspring from year 1 to 4

    Stage 2: offspring from year 4 to 40

    Stage 3: adults\
    pop_transition_by_age —-\> transition happens at max age for each stage (useful when juveniles require decades before transitioning to a reproductive stage

2.  stage survival from COMPADRE matrix, fecundity value (1000) and S1 (survival of Stage1) were previously tested in Stefaninin et al., in prep (previous paper). S3 (adult survival also tested but didn't give differences in the output).

    pop_transition_matrix {{0, 0, 0, 1000}\
    {0.3, 0, 0, 0}\
    {0, 0.7, 0, 0}\
    {0, 0, 1, 0.97}}

3.  (patch number and grid: 25 x 25 grid, each patch is 16\^2 m = 100m x 100m = 1 ha)

4.  patch carrying capacity (patch_nbfem) fixed at 50 (?), from literature:

    -   50 in Waldi (Stefanini et al., in review),
    -   800 in Allenwiller (Stefanini et al., in review),\
    -   roughly 16-48 in favorable conditions in Dobrovolný et al., 2010 (https://jfs.agriculturejournals.cz/pdfs/jfs/2010/12/04.pdf)
    -   max found 8750/ha (14/patch) (but very variable depending on the conditions) in Axer 2021 (https://link.springer.com/article/10.1007/s10342-021-01377-w)
    -   more than 10000/ha (16/patch) in one study in Poland (10.3161/15052249PJE2015.63.2.009)
    -   Gratzler et al., 2022 reported about 630 beech seedlings in 1 ha plot
    -   

---\> 8-16 seedlings = moderate regeneration scenario, 50 seedlings = high regeneration scenario

4.  seed dispersal distance and shape parameter fixed at 10 or 15 m, and \~1 (shape)
    -   29m and 0.5 shape in Waldi (too large in Allenwiller)
    -   7m and 1.2 in Westergren 2023
    -   10m in Oddou-Muratorio 2010
    -   up to 20-25 m the max amount, max at 1-5 m (+ refs in Dobrovolny 2016 https://iforest.sisef.org/contents/?id=ifor1581-008)
    -   20 m radius for the seed falling according to (Wagner S (1999) The initial phase of natural regeneration in mixed ash-beech stands—ecological aspects. Sauerländer)
    -   6m highest in Kutter M, Gratzer G (2006) https://www.researchgate.net/publication/286327613_Current_methods_for_estimation_of_seed_dispersal_of_trees\_-\_An_example_of_seed_dispersal_of_Picea_abies_Abies_alba_and_Fagus_sylvatica);
    -   13-20m in Millerón et al. 2013 (https://doi.org/10.1111/mec.12200 ) --\> other references in dicussion
5.  pollen dispersal distance and pollen shape parameter: 40/50? (need to create spatial structure)
    -   800-1000m and 0.17 in Westergren 2023
    -   41.63m in Oddou-Muratorio 2010
    -   110m and 0.85 in Waldi (previous paper - not converging in Allenwiller)
6.  selfing ( *mating_system 4, + mating_proportion* ): 0.003 from literature
    -   0.0073 in Allenwiller
    -   0.0032 in Wäldi
    -   0.00036 in Westergren 2023
7.  competition model and competition coefficient (regulation_by_competition): 0.005 or 0.01
    -   0.005 in Allenwiller
    -   0.01 in Wäldi

NB: to parametrise the scenario, start from the number of adults in a 1 ha plot and the number of seedlings reported: - Pauline reported a max basal area of approx 0.2-0.3 m\^2 per tree, ---\> use to calcaulte the b coefficient - Allenwiller = 68 adults per ha, Waldi = 34 adults per ha - Gratzler et al., 2022 reported about 630 beech seedlings in 1 ha plot

Testing parameters:

1.  number of trees of relative species –\> 3 values from the questionnaires?
2.  position of trees of relative species –\> **mixed, stripes, clusters**
3.  fitness reduction / selection against orientalis/sylvatica/hybrids? (environmental effect?)

### Spatial architecture (simulation grid)

First we need to define the starting grid:

```{r}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(stringr)
  library(scales)
  library(viridis)
})
```

```{r}
#| label: creating grid
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"

dir.create("C:/Users/stefanin/Dropbox/WSL_PhD/Projects/Hybridization2/input_files")
setwd("C:/Users/stefanin/Dropbox/WSL_PhD/Projects/Hybridization2/input_files")

library(terra)

ext <- ext(0, 100, 0, 100)
r <- rast(
  extent = ext,
  resolution = 4,   # patch size: 4 x 4 meters
  crs = "EPSG:3035" 
)
ncol(r)
nrow(r)
grid_poly <- as.polygons(r)
grid_poly$patch_id <- 1:nrow(grid_poly) ## add patch ID

## number of patchs
length(grid_poly$patch_id)

writeVector(grid_poly, "input_files/grid_4x4m_100x100m.shp", filetype = "ESRI Shapefile",overwrite = TRUE)

```

### configuration files and source_pop

For each spatial configuration, create a master file with the structure as below:

| simulation | patchID | patch_value |
|------------|:--------|------------:|
| beech0     | 1       |          NA |
| beech0     | 2       |           S |
| beech0     | 3       |           O |
| beech0     | 4       |           X |

Each patch can have the following state:

S European beech ---\> quanti = 0, patch_nbfem \> 0

O Oriental beech ---\> quanti = 1, patch_nbfem \> 0

A Abies ---\> quanti = 0, patch_nbfem = 0

X Cut / unavailable ---\> quanti = 0, patch_nbfem = 0

NA Empty land ---\> quanti = 0, patch_nbfem = 0

**NB: in Nemo, patches are arranged row-wise on the grid (see the manual)**

```         
  source_pop          path of the file containing the population loaded to set the starting generation of a replicate, which contain the age class (offspring and/or adults). 
  source_file_type    .txt  =  extension of the source_pop file
  source_preserve     1    = individuals are sequentially imported, and kept (i.e. adults)
  source_fill_stage   [offspring, adults]  for the simulations starting with offspring (empty), or the ones starting with adults.. ?
  source_generation   1   = generaiton to load from the **binary** source file (maybe not needed)
  
  If the source population does not contain enough individuals within patches to fill the receiving patches to the carrying capacity, the landscape is not filled (== maybe I dont need to set carrying capacity 0 for non-beech starting patches??)
  
```

### Quanti trait (species identity)

The species identity is encoded as quantitative trait (quanti) based on loci with 2 possible "alleles":

-   European beech allele = 0
-   Oriental beech allele = 1
-   Hybrids = intermediate values

The input *quanti_init_freq* controls the starting species composition. It consists of a matrix with one value per patch where sylvatica = 0 as default value (empty patches are defined later by the starting carrying capacity matrix, and they will be ignored automatically, but they need to appear in the matrix).

The script uses the write.nemo.matrix to convert the quanti_init_freq matrix into the format which is ready for Nemo. The whole script is looped over all the configurations file inside the *configs_txt* folder: the restult is one quanti_init_file for each configuration file.

```{r}
#| label: quanti_init_file
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"

library(data.table)
library(terra)
dir.create("input_files/quanti_init_freq")
files <- list.files("input_files/configs_txt")
grid <-  vect("input_files/grid_4x4m_100x100m.shp")

# function to write matrix in NEMO format 
write.matrix.nemo <- function(mat, outfile) {
  rows <- dim(mat)[1]
  cols <- dim(mat)[2]
  cat("{", file = outfile)
  
  for(i in 1:rows) {
    cat("{", file = outfile, append = TRUE)
    cat(mat[i,], sep = ",", file = outfile, append = TRUE)
    cat("}\n", file = outfile, append = TRUE)
  }
  
  cat("}\n", file = outfile, append = TRUE)
}

### create a quanti_ini file according to the configuration file with 0 = sylvatica patches and 1 = orientalis patches
for (file in files) {
      
    # importing simulated population with species information and patch ID (check the path)
    sim_name <-  sub("^(.*)\\.txt$", "\\1", file)
    indiv <- fread(paste0("input_files/configs_txt/", file))
    # creating matrix with 0 as default value (empty patches will be ignored automatically, but they need to appear in the matrix)
    quanti_init_freq = matrix(0, ncol = 1, nrow = ncell(grid))
    # assign patch value based on patch id in individuals file
    indiv$quanti_code = ifelse(indiv$patch_value=="O",1,0)
    indiv[is.na(indiv$quanti_code), "quanti_code"] <- 0
    quanti_init_freq[ c(indiv$patchID), ] = indiv$quanti_code
    # write quanti_init matrix to file (or whether you want to store them) and copy to the nemo folder 
    write.matrix.nemo(quanti_init_freq, paste0("input_files/quanti_init_freq/quanti_init_file_",sim_name,".txt"))

}

## creating the initial empty matrix for empty simulations: 
sim_name <- "empty0"
quanti_init_freq = matrix(0, ncol = 1, nrow = ncell(grid))
write.matrix.nemo(quanti_init_freq, paste0("input_files/quanti_init_freq/quanti_init_file_",sim_name,".txt"))

```

### patch_init_stage_size

To set the introduction of individuals in a specific stage (seedlings of 2/3 years old instead of sawing) Matrix where each row = patch, each value = number of individuals in that stage.

E.g. patch_init_stage_size {{0, 20, 0,0}} = 0 Stage 0 , 20 stage1, 0 stage 2, 0 stage 3.

This depends on **how many** seedlings the forester/researcher would set and **at which age** for each configuration. This information is in the questionnaires_all.xlsx file (as the harvesting scheme). Most of the questionnaires suggest planting via direct seeding, or planting seedlings of 2/3 years old. Some questionnaires do not suggest any planting age - in this case we assume direct seeding.

| Direct seeding | 1 yo seedlings | 2-3 yo seedlings | 10 yo seedlings |
|----------------|----------------|------------------|-----------------|
| empty3,7       | beech7         | empty1,4,5,6,8,9 | beech6          |
| mixed6         |                | beech1,2,3,5,8   | mixed3          |
|                |                | mixed1,2,5       |                 |

: Suggested planted methods

NB: this differs for the three sets of simulations

1.  empty land scenario: planting both where patch = S (syl) and O (ori), because nothing is present.
2.  beech and mixed scenario: planting only where patch = O (ori), because S is already present.

NB: for the initial scenario (empty0, beech0, mixed0) there is no need for this

```{r}
#| label: patch_init_stage_size
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"

library(data.table)
library(terra)
dir.create("input_files/patch_init_stage_size")
files <- list.files("input_files/configs_txt")
grid <-  vect("input_files/grid_4x4m_100x100m.shp")

### for each configuration, set the number of planted seedlings and the stage from the table above
planting_scheme <- list(
  direct = c(50, 0, 0, 0),   # direct seeding 20-25 seeds
  age14   = c(0, 20, 0, 0),  # planting 20-25 young seedlings per patch
  age440  = c(0, 0, 1, 0),  # planting one juvenile tree between 4 and 40 yo
  age40  = c(0, 0, 0, 1)   # planting of 1 adult per patch
)

## define which populations have which starting age
# burnin simulations start with adults
planting_method <- list(
  direct = c("empty3","empty7","mixed6"),
  age14  = c("empty1","empty4","empty5","empty6","empty8","empty9",
             "beech1","beech2","beech3","beech5","beech7","beech8",
             "mixed1","mixed2","mixed5"),
  age440  = c("beech6","mixed3"), 
  age40 = c("beech0", "mixed0")
)

# which patch types allow planting for each simulation
planting_patches <- list(
  empty  = c("S", "O"),
  beech  = c("O"),
  mixed  = c("O"),

  # burn-in patches to be read
  beech0 = c("S", "A"),
  mixed0 = c("S", "A")
)

## helper function
get_sim_set <- function(sim_name) {
  if (grepl("^empty", sim_name)) return("empty")
  if (grepl("^beech", sim_name)) return("beech")
  if (grepl("^mixed", sim_name)) return("mixed")
  stop("Unknown simulation set")
}

## function to get patches to plant
get_plantable_patches <- function(sim_name) {
  # exact match first (for beech0 / mixed0)
  if (sim_name %in% names(planting_patches)) {
    return(planting_patches[[sim_name]])
  }

  # otherwise fall back to sim set
  sim_set <- get_sim_set(sim_name)
  planting_patches[[sim_set]]
}

## assign planting mthod to sim name
assign_planting <- function(sim_name) {
  for (m in names(planting_method)) {
    if (sim_name %in% planting_method[[m]]) {
      return(planting_scheme[[m]])
    }
  }
  return(planting_scheme$direct)
}
  
make_patch_init_stage_size <- function(conf, sim_name) {

  n_patches <- nrow(grid)
  mat <- matrix(0, nrow = n_patches, ncol = 4)
  colnames(mat) <- c("stage0","stage1","stage2","stage3")

  scheme <- assign_planting(sim_name)
  plantable_patches <- get_plantable_patches(sim_name)

  for (i in seq_len(n_patches)) {
    ptype <- conf$patch_value[i]

    if (!is.na(ptype) && ptype %in% plantable_patches) {
      mat[i, ] <- scheme
    }
  }

  mat
}

## write in NEMO format
write.matrix.nemo <- function(mat, outfile) {
  rows <- dim(mat)[1]
  cols <- dim(mat)[2]
  cat("{", file = outfile)
  
  for(i in 1:rows) {
    cat("{", file = outfile, append = TRUE)
    cat(mat[i,], sep = ",", file = outfile, append = TRUE)
    cat("}\n", file = outfile, append = TRUE)
  }
  
  cat("}\n", file = outfile, append = TRUE)
}


for (conf_file in files) {

  # derive simulation name correctly
  sim_name <- tools::file_path_sans_ext(conf_file)

  # read the configuration file
  conf <- fread(file.path("input_files/configs_txt", conf_file))

  # build patch_init_stage_size from the *config*, not the grid
  mat <- make_patch_init_stage_size(conf, sim_name)

  # output file
  out_file <- file.path(
    "input_files/patch_init_stage_size",
    paste0("patch_init_stage_size_", sim_name, ".txt")
  )

  write.matrix.nemo(mat, out_file)
}

```

### patch_nbfem (carrying capacity)

The carrying capacity of each simulation can be specified using **patch_nbfem**, which is the maximum number of females that can occupy a patch, but it does not track whether females are currently present (if 0, it means no recruitment in that patch). In other words, it defines where trees can exist, not where they currently exist.

1.  as a single number valid for all the patches: individuals grow homogeneously across the landscape

patch_number 625

patch_nbfem 80 100 300 (3 sequential parameters to test)

patch_nbmal 0

2.  as a nemo-matrix, with the carrying capacity value for each patch: individuals can grow heterogeneously across the landscape. Some patches can have carrying capacity = 0, for instance if they are occupied by other tree species.

patch_number 625\
patch_nbfem &patch_nbfem_80.txt &patch_nbfem_100.txt &patch_nbfem_300.txt (3 sequential parameters to test)

patch_nbmal 0

**NB: patch capacity can vary across time, if set as sequential parameter:**

patch_capacity (\@g0 10, \@g50 100, \@g100 500)

= capacity is 10 at generation 0, increases to 100 at generation 50 and to 500 at generation 100

For the simulations sets *empty* and *beech*, the carrying capacity can be set as homogenous, as the whole grid is available for colonization.

For the *mixed* configurations, a matrix *patch_nbfem* has to be created to set the carrying capacity = 0 where other species are present. We need only one matrix (mixed0) since we assume that the other species composition is the forest will not change over time. However, for the same configuration we can test different values of carrying capacity (the same tested in the previous sets of simulations).

```{r}
#| label: patch_nbfem
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"

library(data.table)
library(terra)
dir.create("input_files/patch_nbfem")

files <- list.files("input_files/configs_txt")
grid <-  vect("input_files/grid_4x4m_100x100m.shp")

# function to write matrix in NEMO format 
write.matrix.nemo <- function(mat, outfile) {
  rows <- dim(mat)[1]
  cols <- dim(mat)[2]
  cat("{", file = outfile)
  
  for(i in 1:rows) {
    cat("{", file = outfile, append = TRUE)
    cat(mat[i,], sep = ",", file = outfile, append = TRUE)
    cat("}\n", file = outfile, append = TRUE)
  }
  
  cat("}\n", file = outfile, append = TRUE)
}

### ONLY FOR ABIES MIXED FOREST: 
# test different values
values <- c(200,500,800)
files_mixed <- files[grep("^mixed0", files)]

for (file in files_mixed) {
  for (value in values) {  
    # importing simulated population with species information and patch ID (check the path)
    sim_name <-  sub("^(.*)\\.txt$", "\\1", file)
    indiv <- fread(paste0("input_files/configs_txt/", file))
    # creating matrix with 0 as default value (empty patches will be ignored automatically, but they need to appear in the matrix)
    patch_nbfem = matrix(0, ncol = 1, nrow = ncell(grid))
    # assign patch value based on patch id in individual files (0 only where abies, otherwise carrying capacity)
    indiv$patch_nbfem = ifelse(indiv$patch_value=="A",0,value)
    indiv[is.na(indiv$patch_nbfem), "patch_nbfem"] <- value
    patch_nbfem[ c(indiv$patchID), ] = indiv$patch_nbfem
    # write quanti_init matrix to file (or whether you want to store them) and copy to the nemo folder 
    write.matrix.nemo(patch_nbfem, paste0("input_files/patch_nbfem/patch_nbfem_",sim_name,"_", value, ".txt"))

  }
}

```

### dispersal (fixed param)

Building Nemo-age seed and pollen dispersal matrices given different

-   mean dispersal distance

-   dispersal kernel shape

-   exponential power function

NB: connectivity / dispersal measure used is accounting also for survival (measure comes from seedlings which were able to germinate and being able to grow)

```{r}
#| label: dispersal input files
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"

library(data.table)
library(terra)
library(ggplot2)
dir.create("input_files/disperse")

# dispersal function (implement such that can be customized)
exp_power_kernel <- function(d, b, x) {
  #  scale parameter a
  a <- d * exp(lgamma(2 / b) - lgamma(3 / b))
  
  #  kernel function K_jk
  return(b * exp(-(x^b / a^b))  / (2* pi * a^2 * exp(lgamma(2/b))))
}

# function to build both reduced dispersal matrices and connectivity matrix
# num_patch = number of patch of the simulation grid
# distance_matrix = pairwise distance matrix between the patches
# d = mean dispersal distance for the kernel
# b = shape of the kernel
# d_thresh = minimum dispersal rate under which dispersal rate is set to 0

build.reduced.dispersal.matrices <- function(num_patch, distance_matrix,  d, b, d_thresh = threshold) {
  stopifnot(dim(distance_matrix)[1] == num_patch)
  
  dispersal_matrix <- matrix(0, nrow = num_patch, ncol = num_patch)
  rate_matrix <- matrix(0, nrow = num_patch, ncol = num_patch)
  connectivity_matrix <- matrix(NA, nrow = num_patch, ncol = num_patch)
  
  for (i in 1:num_patch) {
    for (j in 1:num_patch) {
      # calculate dispersal probability from patch i to patch j
      rate_matrix[i, j] <- exp_power_kernel(d, b, distance_matrix[i, j])
    }
  }
  # define full disp matrix 
  dispersal_matrix <- rate_matrix
  
  for (i in 1:num_patch) {
    # Get indices of the patches sorted by dispersal probability in descending order
    ord <- order(rate_matrix[i, ], decreasing = TRUE)
    
    # re-order the values in row i of reduced rate matrix, connectivity matrix
    rate_matrix[i, ] <- rate_matrix[i, ord]
    connectivity_matrix[i, ] <- ord

    # remove elements below the threshold by setting them equal to NA 
    to_remove <- which(rate_matrix[i, ] < d_thresh)
    rate_matrix[i, to_remove] <- NA
    connectivity_matrix[i, to_remove] <- NA
    
    # normalize dispersal rates to sum to 1 ---> DENSITY KERNEL
    # =each row is a multinomial probability vector
    # rate_matrix[i,j] = probability that one seed from patch i lands in patch j
    sum_d <- sum(rate_matrix[i, ], na.rm = TRUE)
    rate_matrix[i, !is.na(rate_matrix[i, ])] <- rate_matrix[i, !is.na(rate_matrix[i, ])] / sum_d
  }
  
  return(list(dispersal_matrix = dispersal_matrix, connectivity_matrix = connectivity_matrix, rate_matrix = rate_matrix))
}

# function to write (general) matrix in NEMO format 
write.matrix.nemo <- function(mat, outfile) {
  rows <- dim(mat)[1]
  cols <- dim(mat)[2]
  cat("{", file = outfile)
  
  for (i in 1:rows) {
    cat("{", file = outfile, append = TRUE)
    cat(mat[i, ], sep = ",", file = outfile, append = TRUE)
    cat("}\n", file = outfile, append = TRUE)
  }
  
  cat("}\n", file = outfile, append = TRUE)
}

## original function for the dipersal matrix 
# write a reduced dispersal matrix, removing elements flagged with NA
# works for both the connectivity and dispersal rate matrices
# mat        : the input matrix (connectivity or rate)
# outfile    : name of the output text file
# discard.NA : if NA values are discarded or not, if not, calls 'write.matrix.nemo'

write.dispersal.matrix = function(mat, outfile, discard.NA=TRUE) {
  
  if(discard.NA) {
    
    cat("{",file=outfile)
    
    # write every row at once, removing NA elements
    for(i in 1:dim(mat)[1]) {
      cat("{",file=outfile, append=TRUE)
      cat(mat[i,which(!is.na(mat[i,]))],sep=",", file=outfile, append=TRUE)
      cat("}\n",file=outfile, append=TRUE)
    }
    cat("}\n",file=outfile, append=TRUE)
    
  } else {
    write.matrix.nemo(mat, file)
  }
  
}

# modified function with connectivity matrix starting always with focal patch, and rate matrix starting with 1 when the row is empty
# to ensure no empty rows
write.red.dispersal.matrix <- function(connectivity_matrix, rate_matrix, conn_file, rate_file) {
  # Write connectivity matrix
  cat("{", file = conn_file)
  for (i in 1:dim(connectivity_matrix)[1]) {
    cat("{", file = conn_file, append = TRUE)
    # Start with the focal patch number
    cat(i, file = conn_file, append = TRUE)
    if (!all(is.na(connectivity_matrix[i, ]))) {
      cat(",", file = conn_file, append = TRUE)
      # Write the non-NA elements of the row, excluding the focal patch number itself
      non_na_elements <- connectivity_matrix[i, which(!is.na(connectivity_matrix[i, ]))]
      non_na_elements <- non_na_elements[non_na_elements != i]
      cat(non_na_elements, sep = ",", file = conn_file, append = TRUE)
    }
    cat("}\n", file = conn_file, append = TRUE)
  }
  cat("}\n", file = conn_file, append = TRUE)
  
  # Write rate matrix
  cat("{", file = rate_file)
  for (i in 1:dim(rate_matrix)[1]) {
    cat("{", file = rate_file, append = TRUE)
    if (all(is.na(rate_matrix[i, ]))) {
      # If the entire row is NA, write 1
      cat(1, file = rate_file, append = TRUE)
    } else {
      # Write only the non-NA elements of the row
      cat(rate_matrix[i, which(!is.na(rate_matrix[i, ]))], sep = ",", file = rate_file, append = TRUE)
    }
    cat("}\n", file = rate_file, append = TRUE)
  }
  cat("}\n", file = rate_file, append = TRUE)
}

#### SEED DISPERSAL MATRIX ####
# importing grid 
grid <- vect("input_files/grid_4x4m_100x100m.shp")

# build distance matrix between each patch (using centroid of the patch)
centroids = centroids(grid)
plot(centroids)
patch_coords = crds(centroids)
colnames(patch_coords) = c("X", "Y")
patch_coords = as.data.frame(patch_coords)  
patch_coords$patch_ID = grid$patch.ID # attach patch ID
distance_matrix = as.matrix(distance(centroids))

# define number of patches
num_patch = ncell(grid)

# ----- SEED dispersal parameters 

## plot kernel 
b <- c(1,1,1) # shape parameter
d <- c(10,15,20)  # mean dispersal distance

## check the curves
x <- 1:100 
results <- data.frame()
for (i in 1:length(d)) {
  d_i <- d[i]
  b_i <- b[i]
  y_values <- exp_power_kernel(d_i, b_i, x)
  temp <- data.frame(x = x, K = y_values, 
                     d = as.factor(d_i), 
                     b = as.factor(b_i))
  results <- rbind(results, temp)
}

ggplot(results, aes(x = x, y = K, group =d, color = d))+
  geom_line(size = 1) +
  labs(x = "Euclidean Distance",y = "Probability of dispersal")

# final parameters
# define threshold for rate going to 0 at tot meters from focal tree
d_t <- 50
d <- 15
b <- 1
threshold = exp_power_kernel(d,b, d_t) ## otherwise, just keep d
threshold

# building matrices and plot
dispersal_matrices = build.reduced.dispersal.matrices(num_patch, distance_matrix, d, b, threshold)

dispersal_matrix <- dispersal_matrices$dispersal_matrix
connectivity_matrix <- dispersal_matrices$connectivity_matrix
rate_matrix <- dispersal_matrices$rate_matrix

## check dimensions
dim(dispersal_matrix)
dim(connectivity_matrix)
dim(rate_matrix)

# check that rate for each row sums to 1
rowSums(rate_matrix, na.rm = TRUE)

## plot how many seeds fall in each patch if a tree is dispersing 1000 seeds
mati=matrix(dispersal_matrix[,num_patch],ncol=n_cols,byrow= T)
mati=mati/sum(mati) # normalization
par(mar = c(4, 4, 4, 10))
plot(mati)
b=reshape2::melt(mati,id.vars="rows")
b$value[b$value*1000<1]=NA ## set <1 seeds as NA
ggplot(data=b,aes(x=Var1,y=Var2,fill=value*1000))+geom_raster()+scale_fill_gradientn(colours = terrain.colors(20),trans = "log")

# write connectivity and rate matrix to files 
# save d = mean dispersal distance of the kernel
write.red.dispersal.matrix(connectivity_matrix, rate_matrix, 
                           paste0("input_files/disperse/Seed_connectivity_matrix_d", d, ".txt"), 
                           paste0("input_files/disperse/Seed_rate_matrix_d", d, ".txt"))

# ----------POLLEN dispersal parameters (here we use Waldi dispersal parameters)
b <- c(1,1,1) # kernel shape
d <- c(40,50,60)  # mean dispersal distance
## check the curves
x <- 1:1000 
results <- data.frame()
for (i in 1:length(d)) {
  d_i <- d[i]
  b_i <- b[i]
  y_values <- exp_power_kernel(d_i, b_i, x)
  temp <- data.frame(x = x, K = y_values, 
                     d = as.factor(d_i), 
                     b = as.factor(b_i))
  results <- rbind(results, temp)
}

ggplot(results, aes(x = x, y = K, group =d, color = d))+
  geom_line(size = 1) +
  labs(x = "Euclidean Distance",y = "Probability of dispersal")


# final parameters
# define threshold for rate going to 0 at tot meters from focal tree
d_t <- 100
d <- 40
b <- 1
threshold = exp_power_kernel(d,b, d_t) ## otherwise, just keep d
threshold

# building matrices and plot
dispersal_matrices = build.reduced.dispersal.matrices(num_patch, distance_matrix, d, b, threshold)

dispersal_matrix <- dispersal_matrices$dispersal_matrix
connectivity_matrix <- dispersal_matrices$connectivity_matrix
rate_matrix <- dispersal_matrices$rate_matrix

## check dimensions
dim(dispersal_matrix)
dim(connectivity_matrix)
dim(rate_matrix)

# check that rate for each row sums to 1
rowSums(rate_matrix, na.rm = TRUE)

## plot how many pollen grain fall in each patch if a tree is releasing 10000 pollens
mati=matrix(dispersal_matrix[,num_patch],ncol=n_cols,byrow= T)
mati=mati/sum(mati) # normalization
par(mar = c(4, 4, 4, 10))
plot(mati)
b=reshape2::melt(mati,id.vars="rows")
b$value[b$value*10000<1]=NA ## set <1 seeds as NA
ggplot(data=b,aes(x=Var1,y=Var2,fill=value*10000))+geom_raster()+scale_fill_gradientn(colours = terrain.colors(20),trans = "log")

# double check distance against dispersal
distance_v <- distance_matrix[lower.tri(distance_matrix, diag = T)][order(row(distance_matrix)[lower.tri(row(distance_matrix), diag = T)])]
dispersal_v <- dispersal_matrix[lower.tri(dispersal_matrix, diag = T)][order(row(dispersal_matrix)[lower.tri(row(dispersal_matrix), diag = T)])]
plot(distance_v, dispersal_v)

# write connectivity and rate matrix to files 
# save d = mean dispersal distance of the kernel
write.red.dispersal.matrix(connectivity_matrix, rate_matrix, 
                           paste0("input_files/disperse/Pollen_connectivity_matrix_d", d, ".txt"), 
                           paste0("input_files/disperse/Pollen_rate_matrix_d", d, ".txt"))
```

### selection

We want to implement potential selection against one of the two pure species, to reproduce potential cases of differential fitness of the pure species and see if there is any case of extinction. We simulated two cases of selection (or differential fitness):

1.  O. beech is more adapted to dry climates: fitness is maximal around phenotype O. beech, while drops as the phenotype moves away from that (hybrids may be favoured or disfavoured):

    selection against E. beech ---\> directional selection implemented by stabilizing selection with the **optimum at + 1 (O. beech)**, with fitness declined smoothly toward P1 (z) = -1 (hybrids have an intermediate value). The selection_fitness_model set the fitness of the individual independent from the fitness of the other individuals of the population (--\> possible population decline, stronger eco-evo feedback).

2.  E. beech is more locally adapted:fitness is maximal around phenotype E. beech, while drops as the phenotype moves away from that (hybrids may be favoured or disfavoured):

    The same as Case 1 but selection_local_optima {{-1.0}} (E. beech is the local optima)

*Implementation:*

Selection variance (w\^2) is the quadratic of the inverse of the selection strenght, therefore, the smaller, the stronger the selection, the larger, the weaker the selection.

If

-   w\^2 = 4, fitness (survival probability) is 0 for European beech = is not surviving,

-   w\^2 = 10, fitness is 0.4 for E beech (60% mortality difference)

-   w\^2 = 20, fitness is 0.8 for E beech (20% mortality difference)

in the ini file:

```         
# LIFE CYCLE EVENTS #

quanti_init 0 
breed_disperse 1 
seed_disperse 2 
viability_selection 3 
regulation 4 
save_stats 5 
aging_multi 6 
save_files 7 

## ----------------------------
## SELECTION
## ----------------------------
viability_selection        3
selection_trait_dimension  1
selection_trait            quant
selection_model            quadratic
selection_fitness_model    absolute
selection_at_stage         {{0}}
selection_local_optima     {{1}}
selection_matrix         {{20}}
```

## Estimating operational cost

In our case, cost differences between simulations will mainly come from two variables: number of introduced seedlings, and operational efficiency of the planting strategy (not the seedling price, which is identical per plant).

Since hard numbers strongly depend on the country, labour cost, terrain and machinery, our strategy is to identify cost components, estimate how the spatial design affects them and reank the designs by expected operational cost.

### Main cost components

The total cost per ha is usually given by:

1.  seedlings cost (cost per seedling \* number of seedlings) —\> increasing with % orientalis
2.  transport —\> we ignore
3.  site preparation —-\> we ignore
4.  planting labor —–\> our focus
5.  protection —–\> we ignore, but is affected by the planting scheme
6.  maintenance ——\> we ignore
7.  supervision and monitoring ——\> we ignore

### Estimating cost of spatial design

1.  dispersed:

    -   Workers move constantly
    -   hard supervision and layout,
    -   maintenance scattered everywhere
    -   protection has to be designed individually (like tree shelters)

    These factors translate into high walking time, no machinery efficiency, no shared fencing

    **Expected cost: highest**

2.  one large cluster:

    -   very efficient planting (minimal movement)
    -   easier site preparation
    -   one single fence for the whole patch
    -   concentrated maintenance

    **Expected cost: lowest**

3.  several small clusters:

    -   intermediate efficiency
    -   some walking time between the clusters
    -   one fencing per cluster
    -   maintenance is partially concentrated

    **Expected cost: intermediate (more than one large cluster, cheaper than fully mixed)**

4.  transects:

    -   very efficient movement
    -   machinery access possible
    -   easy maintenance along lines
    -   protection is likely individual (unless fenced strips)

    **Expected cost: lower to intermediate (more than one large cluster, cheaper than fully mixed)**

Summary - spatial aggregation reduces operational cost, therefore the likely cost ranking is

1.  one cluster
2.  transects
3.  multi cluster
4.  completely mixed

To combine this ranking with the proportion of orientalis introduced, we applied this strategy: we assigned a cost multiplier (something like "relative cost index") to the configurations, and estimated only the average cost (forestry price, not garden center price) of 2 euros per plant multiplied by the number of plants introduced (ori_prop_cost). The final estimated baseline cost was estimated by multiplying these two values.

For each proportion of orientalis introduced, the number of planted seedlings is:

-   10 % patches: 62 patches \* 20 seedlings/patch = 1240 seedlings \* 2 = E

-   25% patches: 156 patches \* 20 seedlings/patch = 3120 seedlings \* 2 = E

-   40% patches: 250 patches \* 20 seedlings/patch = 5000 seedlings \* 2 = E

Which multiplied by the cost index give the cost of all the scenarios:

```{r}
#| label: cost estimation
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"

cost_per_seedling <- 2

seedlings <- data.frame(ori_prop = c("10%", "25%", "40%"),
                        n_seedlings = c(1240, 3120, 5000))

cost_index <- data.frame(design = c("one_cluster","transects","multi_cluster","completely_mixed"),
  cost_multiplier = c(1.00, 1.10,1.25,1.50))

# baseline cost (seedlings only)
seedlings$baseline_cost <- seedlings$n_seedlings * cost_per_seedling

# combine all and order by cost
final_cost_table <- expand_grid(seedlings, cost_index) %>%
  mutate(estimated_cost = baseline_cost * cost_multiplier) %>%
  dplyr::arrange(ori_prop, cost_multiplier)
final_cost_table

write.csv(final_cost_table,"C:/Users/stefanin/Dropbox/WSL_PhD/Projects/Hybridization2/Cost_design_table.csv" )

```

This cost estimate should be attached to the final results combined table as a new variable.

## Analyse results of sensitivity analysis

For this step we want to check:

1.  if the hybrid proportions and distribution are consistent across replicates of the same simulation

2.  if the hybrid proportions and distribution are consistent across different values of carrying capacity (K) and competition coefficient (B).

For each scenario:

1.  **Hybrid proportions** at generation X for each life stage

2.  **Hybrid patterns across the grid/spatial extent of introgression**

### Demographic patterns

First checking the demographic trends across the years for the different parameter combinations.

```{r}
#| label: demographic results
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"

library(ggplot2)
library(dplyr)
library(tidyr)
library(raster)
library(terra)
library(data.table)
library(viridis)
library(readr)
library(stringr)
library(ggh4x)
library(sf)
library(xlsx)
library(tidyverse)
library(ggpubr)

### NB FIRST REMOVE THE EXISTING RESULTS

## move the files on Windows to analyse
res_path  <- "//wsl.localhost/Ubuntu/home/camilla/nemoage0.32.6b_2/exe/results/"
res_path2 <- "C:/Users/stefanin/Dropbox/WSL_PhD/Projects/Hybridization2/Results/"
dir.create(res_path2, recursive = TRUE, showWarnings = FALSE)
file.copy(from = list.files(res_path, full.names = TRUE), to=res_path2, recursive = TRUE)

# OR
# res_path2 <- "C:/Users/stefanin/Downloads/WSL-sylvatica-orientalis-hybridization-main/WSL-sylvatica-orientalis-hybridization-main"

####---------- demographic trend per simulation

# files with demographic data
file_pattern <- "sc_p20\\.txt$" 
file_list <- list.files(path = res_path2, recursive = TRUE, full.names = TRUE, include.dirs = TRUE, pattern = file_pattern)
file_list

# define which columns you want to read from the files
cols_to_read = c("replicate", "generation", "pop.tot", "a0.tot", "a1.tot","a2.tot", "a3.tot")

# read each file only the selected columns and label with name and create dataframe for plot 
library(data.table)

data_list <- vector("list", length(file_list))
names(data_list) <- basename(file_list)

for (i in seq_along(file_list)) {
  file <- file_list[i]
  data <- fread(file, select = cols_to_read)
  
  # extract file name
  file_name <- gsub(".txt", "", basename(file))
  # extract the type of simulation
  sim_name <- sub("_.*$", "", file_name)
  
  k = as.numeric(sub(".*_k([0-9.]+).*", "\\1", file_name))
  b = as.numeric(sub(".*_b([0-9.]+).*", "\\1", file_name))
 
  data$sim_name <- sim_name
  data$k <- k 
  data$b <- b
  
  pop_data = melt(data, id.vars = c("replicate", "generation", "sim_name", "k", "b"), 
                  measure.vars = patterns(".tot"), 
                  variable.name = "stage", value.name = "N_stage")
  
 data_list[[i]] <- pop_data
}

# cobine all simulations
combined_pop_data <- rbindlist(data_list, use.names = TRUE)

# save
saveRDS(combined_pop_data,file = file.path(res_path2, "Demographic_data_per_simulation.RDS"))
rm(data_list, data, pop_data)

##----- summarize data per patch
# importing grid 
grid <- vect("input_files/grid_4x4m_100x100m.shp")
grid_r <- terra::rast(ext = ext(grid), resolution = 4, crs = "EPSG:3035")
# grid dimensions 
n_rows = dim(grid_r)[1]
n_cols = dim(grid_r)[2]

# Stages (NB: some old versions files a0 columns are instead called "off")
stages <- c("a0", "a1", "a2", "a3")

# get the patch data from the filelist
data_list2 = list()

for (i in seq_along(file_list)) {
  file <- file_list[i]
  
  tryCatch({
    data <- fread(file)
    
    # extract file name
    file_name <- gsub(".txt", "", basename(file))
    # extract the type of simulation
    sim_name <- sub("_.*$", "", file_name)
    
    k = as.numeric(sub(".*_k([0-9.]+).*", "\\1", file_name))
    b = as.numeric(sub(".*_b([0-9.]+).*", "\\1", file_name))

    
    #rename `off.` columns to `a0.`
    colnames(data) <- gsub("^off\\.", "a0.", colnames(data))

    data$sim_name <- sim_name
    data$k <- k 
    data$b <- b

    # melt data for each stage 
    stages_data_list <- lapply(stages, function(stage) {
      # Process females
      fem_data <- melt(data, 
                       id.vars = c("replicate", "generation", "sim_name", "k", "b"), 
                       measure.vars = grep(paste0("^", stage, "\\.fem\\.p\\d+$"), colnames(data), value = TRUE), 
                       variable.name = "patch", 
                       value.name = "N_ind")
      
      # patch number from column name
      fem_data[, patch := as.numeric(gsub(paste0(stage, ".fem.p"), "", patch, fixed = TRUE))]
      
      # stage column
      fem_data$stage <- stage
      
      # grid information
      fem_data$n_rows <- n_rows
      fem_data$n_cols <- n_cols
      
      return(fem_data)
    })
    
    # combine stage data
    file_data <- rbindlist(stages_data_list)
    
    # store the processed file with a unique identifier (avoid overwriting)
    # extract file name
    file_name <- gsub(".txt", "", basename(file))
    data_list2[[paste0(file_name, "_", i)]] <- file_data
    
  }, error = function(e) {
    # Catch and print any errors during file processing
    cat("Error processing file:", file, "\n")
    print(e)
  })
}

# combine in 1 df
combined_patch_data = do.call(rbind, data_list2)
saveRDS(combined_patch_data,file = file.path(res_path2, "Demographic_data_per_patch.RDS"))
rm(data, file_data, data_list2, stages_data_list)

```

```{r}
#| label: demographic results plot2
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"

library(ggplot2)
library(dplyr)
library(tidyr)
library(raster)
library(terra)
library(data.table)
library(viridis)
library(readr)
library(stringr)
library(ggh4x)
library(sf)
library(xlsx)
library(tidyverse)
library(ggpubr)

res_path2 <- "C:/Users/stefanin/Dropbox/WSL_PhD/Projects/Hybridization2/Results/"
# load df
combined_pop_data <- readRDS(file.path(res_path2, "Demographic_data_per_simulation.RDS"))
combined_patch_data <- readRDS(file.path(res_path2, "Demographic_data_per_patch.RDS"))

# importing grid 
grid <- vect("input_files/grid_4x4m_100x100m.shp")
grid_r <- terra::rast(ext = ext(grid), resolution = 4, crs = "EPSG:3035")
# grid dimensions 
n_rows = dim(grid_r)[1]
n_cols = dim(grid_r)[2]

# add label for stage
nemo_stage <- c("pop.tot" = "Total Population","a0.tot" = "Stage 0", "a1.tot" = "Stage 1", "a2.tot" = "Stage 2", "a3.tot" = "Stage 3")
combined_pop_data$nemo_stage <- nemo_stage[combined_pop_data$stage]
# check n replicates
combined_pop_data %>%group_by(sim_name,generation,k,b, stage) %>% count()

# ----- demographic trends average simulation
ggplot(combined_pop_data, aes(x = generation, y = N_stage, group = interaction(nemo_stage, replicate), col = nemo_stage)) +
      geom_line(linewidth = 0.2) +
      facet_grid(k ~ sim_name+b, scales="free_y") +
      labs( x = "Generations", y = "Number of individuals",
        color = "Nemo Stage"
      ) +
      scale_y_continuous(labels = scales::comma) +
      theme_bw()


## -- demographic trends across the grid
# add label for stage
nemo_stage <- c("a0" = "Stage 0", "a1" = "Stage 1", "a2" = "Stage 2","a3" = "Stage 3")
combined_patch_data$nemo_stage <- nemo_stage[combined_patch_data$stage]

# subset only generations that are multiples of 20 and remove after 200 (population is stable)
subset_data <- combined_patch_data[combined_patch_data$generation == 1 | combined_patch_data$generation %% 200 == 0, ]

# mean number of individuals per patch across replicates
mean_patch_data = subset_data %>%
  group_by(nemo_stage, sim_name, generation, patch, n_rows, n_cols, k,b) %>%
  summarize(mean_N_ind = mean(as.numeric(N_ind), na.rm = TRUE)) %>%
  ungroup()

##---> subset for the nemostage
stage = "Stage 3"

ggplot(subset(mean_patch_data, nemo_stage == stage), aes(x = (patch - 1) %% n_rows + 1,   y = n_cols - ((patch - 1) %/% n_cols + 1),  fill = mean_N_ind)) +
  geom_tile() +
  # scale_fill_viridis(trans = "log") +
  scale_fill_gradientn(colors = c("grey","blue", "purple","red", "orange","gold", "yellow", "green"), limits = c(0, 20),
                       breaks = seq(0, 20, by = 5)) +
  facet_grid( k + b  ~  sim_name+ generation ) +
  labs(title = stage,x = "", y = "", fill = "Number of individuals / patch") +
  theme_void()+
  theme(aspect.ratio = 1, legend.position = "bottom")
```

### Species pattern (quanti trait)

First check to the distribution of the quanti trait, i.e. the species identity. It is based on the quantitative trait P1, where

1 = pure *orientalis*

-1 = pure *sylvatica*

0 = F1 hybrid

intermediate = advanced generation hybrids

**NB: MAKE SURE THAT THE FILENAME METADATA ARE CONSISTENT ACROSS SIMULATIONS, OTHERWISE THE FILES ARE NOT PROCESSED CORRECTLY**

```{r}
#| label: processing quanti results
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"

# the output is a unique file : Quanti_data_Allenwiller_year2020.RDS
res_path2 <- "C:/Users/stefanin/Dropbox/WSL_PhD/Projects/Hybridization2/Results/"
# res_path2 <- "C:/Users/stefanin/Downloads/WSL-sylvatica-orientalis-hybridization-main/WSL-sylvatica-orientalis-hybridization-main"

# to select all the replicates
file_pattern <- ".*\\.quanti$"

# to select only 30 replicates
#file_pattern <- ".*_results_([0-9]{2,3})_((0[1-9])|(1[0-9])|(2[0-9])|(30))\\.quanti$"

quanti_files <- list.files(path = res_path2, recursive = TRUE,full.names = TRUE,include.dirs = TRUE,pattern = file_pattern)

quanti_list <- vector("list", length(quanti_files))

for (i in seq_along(quanti_files)) {

  file <- quanti_files[i]
  data <- fread(file)
  ori_proportion <- as.numeric( stringr::str_extract(basename(file), "(?<=_p)\\d+"))
  configuration <-str_match(basename(file),"_p\\d+_(.*?)_rep")[, 2]
  k <- as.numeric(sub(".*_k([0-9.]+).*", "\\1", file))
  b <- as.numeric(sub(".*_b([0-9.]+).*", "\\1", file))

 # generation-- any 3-digit number surrounded by _xxx_
  generation <- as.numeric( str_extract(basename(file), "(?<=_)\\d{3}(?=_)"))
 # replicate = last number before .quanti
  replicate <- as.numeric(str_extract(basename(file), "(?<=_)\\d+(?=\\.quanti$)"))

  data[, `:=`(
    configuration  = configuration,
    ori_proportion = ori_proportion,
    k         = k,
    b         = b,
    generation = generation,
    replicate  = replicate
  )]

  quanti_list[[i]] <- data
}

# Combine all files into one data.table ad save
quanti_data_all <- rbindlist(quanti_list, fill = TRUE)
# saveRDS(quanti_data_all,file = file.path(res_path2, "Quanti_data.RDS"))
#rm(quanti_files, quanti_list)

```

For each simulation and replicate, plot the mean p1 value (species identity) across all the individuals of the patch (to have a very general overview of the result of the simulation):

*NB: each line is a replicate*

```{r}
#| label: plot P1 across time
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"


quanti_data_all <- readRDS(file.path(res_path2, "Quanti_data.RDS"))

## summarise data per patch = each patch has the average P1 value across individuals of the patch
quanti_average <- quanti_data_all %>%
  group_by(configuration, ori_proportion, replicate, generation, stage, k,b) %>%
  summarize(P1 = mean(P1, na.rm = T), .groups = "drop")

ggplot(quanti_average,aes(x = as.numeric(generation), y = as.numeric(P1),group = interaction(replicate, factor(stage)),  color = factor(stage))) +
    geom_line() +
    #scale_fill_viridis_c(option = "viridis", direction = -1) +
    facet_grid( ori_proportion+ b~ configuration +k) + 
    labs(x = "Generations", y = "Average P1 across the whole stand") +
    theme_bw()+
    theme(
      aspect.ratio = 1,
      legend.position = "right",
      panel.spacing.x = unit(1, "lines"),
      panel.spacing.y = unit(0.1, "lines"),
      strip.text = element_text(size = 10))

```

Checking if results are consistent across replicates using the Coefficient of Variation (CV, SD / mean). As SD increases with the mean, you cannot compare the mean across simulations that have different mean, so CV does not depend on the mean anymore since it is divided by the mean. It shows the variation relative to the mean. The higher the CV, the higher the SD relative to the mean.

```{r}
#| label: hybrid proportions replicates check
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"


# caluclate the proportion of hybrids for each simulation (data table is faster)
library(data.table)

dt <- copy(quanti_data_all)
generation <- 200

hyb_prop <- dt[generation == generation,
  .( N = .N,N_hybrid = sum(P1 > -1 & P1 < 1),prop_hybrid = mean(P1 > -1 & P1 < 1)), by = .(configuration, ori_proportion, replicate, stage, k, b) ]

## calculate mean sd and cv per sim
replicate_cv <- hyb_prop %>%
  group_by(configuration, ori_proportion, stage, k, b) %>%
  summarize(
    mean_hi = mean(prop_hybrid),
    sd_hi   = sd(prop_hybrid),
    cv_hi   = sd_hi / mean_hi,
    .groups = "drop"
  )


## check the sd
ggplot(replicate_cv,
       aes(x = factor(stage), y = sd_hi, fill = factor(stage))) +
  geom_boxplot() +
  facet_grid(ori_proportion + b ~ configuration + k) +
  labs(
    y = "SD of hybrid proportion",
    x = "Stage"
  ) +
  theme_bw()

## check the CV
ggplot(replicate_cv,
       aes(x = factor(stage), y = cv_hi, fill = factor(stage))) +
  geom_boxplot() +
  facet_grid(ori_proportion + b ~ configuration + k) +
  labs(
    x = "Stage",
    y = "Coefficient of variation (CV)",
    title = "Replicate variability across simulation parameter combinations"
  ) +
  theme_bw()


```

Now plot the hybrid proportions across the whole stand for each simulation, age class (stage) and at generation X.

```{r}
#| label: hybrid proporiton boxplot
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"


## boxplots for each simulation with x = stage, y = prop, color = hybrid class
ggplot(hyb_prop, aes(x = stage, y = prop_hybrid, group = factor(stage), fill = factor(stage)))+
  geom_boxplot()+ theme_bw()+  
  facet_grid( ori_proportion+ b~ configuration +k) 


```

------------------------------------------------------------------------

Next question: **within the same configuration and proportion of orientalis, do K and B significantly change the hybrid proportions?**

Basically, we want to see if the main results are robust across the K and B values. To do that, we

1.  run the simulations with the combinations of K and B
2.  plot our main figure x = configuration_oriproportion; y = hybrid_proportions and facet_grid (create different panels) for K and B. We compare the distirbutions of the data with statistical test (ANOVA).

--\> are the results consistent? E.g. is the hybrid proportion consistently highest(lowest) in the same configuration x ori_proportion across all k and b values, or is it changing? If the main results are changing for different K and B values, the significantly affect the hybridization.

```{r}
#| label: hybrid proporiton boxplot 2
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"


## select only one stage (e.g. adults)
hyb_prop_adults <- subset(hyb_prop, stage == 2)

## boxplots for each simulation with x = configuration, y = prop, color = % orientalis
ggplot(hyb_prop_adults, aes(x = configuration, y = prop_hybrid, group = interaction(configuration, ori_proportion), fill = factor(ori_proportion)))+
  geom_boxplot()+ theme_bw()+  
  facet_grid( b~ k) 


```

------------------------------------------------------------------------

## Analyse results of final simulations

Recap of the dimensions of our results: we have

-   3 selection scenarios (neutral, selection against one of the Pure species)

-   4 planting designs/configurations

-   3 proportions of introduced species

-   3 age classes (stage 0 is not considered and likely not even stored by Nemo)

For each selection scenario, we want to assess which planting design maximizes hybridization and its persistence through time, considering each age class.

Expectation: we expect that, regardless the selection scenario

1.  a higher proportion of introduced species will increase hybridization
2.  a more mixed design will increase hybridization

We will assess the short-term hybridization (2 or 10 years?), mid-term hybridization (100 years?) and long term hybridization (500 or 1000 years?).

```{r}
#| label: processing quanti results 2
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"

library(data.table)
library(stringr)

# the output is a unique file : Quanti_data_Allenwiller_year2020.RDS
res_path2 <- "C:/Users/stefanin/Dropbox/WSL_PhD/Projects/Hybridization2/Results/"
# res_path2 <- "C:/Users/stefanin/Downloads/WSL-sylvatica-orientalis-hybridization-main/WSL-sylvatica-orientalis-hybridization-main"

# to select all the replicates
file_pattern <- ".*\\.quanti$"

# to select only 30 replicates
#file_pattern <- ".*_results_([0-9]{2,3})_((0[1-9])|(1[0-9])|(2[0-9])|(30))\\.quanti$"

quanti_files <- list.files(path = res_path2, recursive = TRUE,full.names = TRUE,include.dirs = TRUE,pattern = file_pattern)

# remove the burnin  files
quanti_files <- quanti_files[!grepl("burnin", quanti_files, ignore.case = TRUE)]

quanti_list <- vector("list", length(quanti_files))

for (i in seq_along(quanti_files)) {

  file <- quanti_files[i]
  data <- fread(file)
  s <- str_match(basename(file), "_r\\d+_(.*?)_k")[,2]
  ori_proportion <- as.numeric( stringr::str_extract(basename(file), "(?<=_p)\\d+"))
  configuration <- str_match(basename(file), "^(.*?)_p")[,2]
  
  #k <- as.numeric(sub(".*_k([0-9.]+).*", "\\1", file)) # -- not needed
  #b <- as.numeric(sub(".*_b([0-9.]+).*", "\\1", file)) # -- not needed
  
   run <- sub(".*\\/(r[0-9]+)\\/.*", "\\1", file)
   # generation-- any 4-digit number surrounded by _xxx_
   generation <- as.numeric( str_extract(basename(file), "(?<=_)\\d{4}(?=_)"))
  # replicate = last number before .quanti
  replicate <- as.numeric(str_extract(basename(file), "(?<=_)\\d+(?=\\.quanti$)"))

  data[, `:=`(
    configuration  = configuration,
    ori_proportion = ori_proportion,
    selection = s,
    #k         = k,
    #b         = b,
    generation = generation,
    run = run,
    replicate  = replicate
  )]

  quanti_list[[i]] <- data
}

#  all files into one data.table 
quanti_data_all <- rbindlist(quanti_list, fill = TRUE)

# remove useless columns
quanti_data_all_clean <- quanti_data_all[, c("sex","home","ped","isMigrant","father","mother") := NULL]

# checks

# assign cost to scenario

# saveRDS(quanti_data_all_clean,file = file.path(res_path2, "Quanti_data.RDS"))
# rm(quanti_files, quanti_list, quanti_data_all, quanti_data_all_clean)
```


### P1 distribution

First we look at the overall distribution of the species phenotype (i.e. P1):

```{r}
#| label: P1 distribution

library(dplyr)
library(ggplot2)

res_path2 <- "C:/Users/stefanin/Dropbox/WSL_PhD/Projects/Hybridization2/Results/"
quanti_data_all <- readRDS(file.path(res_path2, "Quanti_data.RDS"))
hist(quanti_data_all$P1)
```

---\> Here we can add Jannis P1 heatmap!

### Overall hybrid proportions and hybrid Index

Then we define a general **hybrid metric which includes all the hybrid classes (F1, F2, BC),** i.e. everything that is not pure.

We also define the Hybrid Index, as the proportion of heterozygous loci across the quanti trait, which ranges from 0 (pure), to 1 (F1) hybrids, with 0.5 and intermediate values representing advanced generation hybrids. We calculated HI as 1 - \|P1\|.

We then check the trajectory over time of the proportion of overall hybrids and mean HI for each simulation, after summarizing across replicates.

```{r}
#| label: calcaulte hyb proportion and HI

# quanti_data_all <- readRDS(file.path(res_path2, "Quanti_data.RDS"))
dt <- copy(quanti_data_all)
rm(quanti_data_all)

# mean hybrid proportions and HI values per simulation and replicate (across all individuals) = 1 row x sim replicate x generation x stage
hyb_prop <- dt[,.( 
  N = .N,
     N_hybrid = sum(P1 > -1 & P1 < 1),
     prop_hybrids = mean(P1 > -1 & P1 < 1), 
     mean_HI = mean(1 - abs(P1))
  ),
     
  by = .(configuration, ori_proportion,selection,generation,run, replicate, stage) ]

gc()

# collapse simulation replicates (average for each run)
hyb_run_mean <- hyb_prop[, .(
  mean_prop = mean(prop_hybrids),
  sd_prop   = sd(prop_hybrids),
  mean_HI = mean(mean_HI),
  sd_HI = sd(mean_HI)
),
by = .(configuration, ori_proportion, selection,run, generation, stage)]


# plot the proportion of hybrids over time across parameters
ggplot(hyb_run_mean,
       aes(generation, mean_prop, colour = factor(stage), group = interaction(factor(stage), factor(run)))) +
  geom_line() +
  geom_ribbon(aes(ymin = mean_prop - sd_prop,
                  ymax = mean_prop + sd_prop,
                  fill = factor(stage)),
              alpha = 0.25,
              colour = NA) +
  facet_grid(configuration ~ selection + ori_proportion) +
  theme_bw()

# plot the HI over time across parameters
ggplot(hyb_run_mean,
       aes(generation, mean_HI, colour = factor(stage), group = interaction(factor(stage), factor(run)))) +
  geom_line() +
  geom_ribbon(aes(ymin = mean_HI - sd_HI,
                  ymax = mean_HI + sd_HI,
                  fill = factor(stage)), alpha = 0.1,  colour = NA) +
  facet_grid(configuration ~ selection + ori_proportion) +
  theme_bw()

```

Subsetting only some specific time points:

```{r}
#| label: comparing overall hybrids proportions for one year

hyb_run_mean_100 <- subset(hyb_run_mean,generation == 100)

# generation = 100

ggplot(hyb_run_mean_100,
       aes(configuration, mean_prop, fill = factor(ori_proportion), group = interaction(configuration, ori_proportion))) +
  geom_boxplot() +
  facet_grid( stage ~ selection) +
  theme_bw()+
  labs(title = "proportion hybrids year 100")

ggplot(hyb_run_mean_100,
       aes(configuration, mean_HI, fill = factor(ori_proportion), group = interaction(configuration, ori_proportion))) +
  geom_boxplot() +
  facet_grid( stage ~ selection) +
  theme_bw()+
  labs(title = "HI year 100")


hyb_run_mean_1000 <- subset(hyb_run_mean,generation == 1000)

# generation = 1000
ggplot(hyb_run_mean_1000,
       aes(configuration, mean_prop, fill = factor(ori_proportion), group = interaction(configuration, ori_proportion))) +
  geom_boxplot() +
  facet_grid( stage ~ selection) +
  theme_bw()+
  labs(title = "proportion hybrids year 1000")

ggplot(hyb_run_mean_1000,
       aes(configuration, mean_HI, fill = factor(ori_proportion), group = interaction(configuration, ori_proportion))) +
  geom_boxplot() +
  facet_grid( stage ~ selection) +
  theme_bw()+
  labs(title = "HI year 1000")


```


For each parameter combination, we ran 5 biological replicates, i.e. 5 different initial planting configuration (random spatial variation) to represent ecological stochasticity in starting conditions (i.e. random spatial draws). Then for each biological replicate, we ran 30 simulation replicate with same starting condition, which represent demographic /stochastic simulation noise. This results in a hierarchical (nested) design.

::: callout-note
We create multiple starting scenarios because spatial initial conditions can matter: even with identical parameter values, slightly different planting configurations may alter local competition, change dispersal dynamics, influence clustering effects, affect early stochastic extinction events.

Using only one starting configuration could create results that are **conditional on that specific spatial realization**. So biological replicates allow to estimate how robust your results are to variation in initial spatial configuration. In ecological simulations, this is often called: “initial condition stochasticity” or “between-realization variance” or “process vs structural variance”.
:::

**Our question is: How much variance is due to initial configuration compared to stochastic simulation noise and compared to parameter effects?**

In this case, an ANOVA test between the biological replicates is not suitable, because they are not treatments, but a random realization of the same treatment. Also, since significance depends on sample size, ANOVA might detect significant small differences just due to the many replicates, without biological relevance.

Instead, biological replicates should be treated as random effect in a mixed model where

`Response ~ param1 * param2 + (1 | biological_replicate)`

This would estimate the variance between biological replicates and the variance within biological replicates. If the between-biological replicates variance is small, results are robust to initial planting, if is large, the initial spatial configuration strongly affects outcomes and is a biological result (we should report it and not eliminate it).

We have the following design:

-   **Fixed factors (treatments)**

    -   `selection`

    -   `configuration`

    -   `ori_proportion`

-   **Time structure**

    -    `generation`

-   **Demographic structure**

    -   `stage` (age class)

-   **Nested stochastic structure**

    -   `run` = biological replicate

    -   `replicate` = simulation replicate (nested within run)

Potentially, summarize the results at the biological replicate level with means (but only 5?).

Evaluating the variance between the runs:

::: callout-note
run      (Intercept) vAR SD == between run Variance and sd

Residual             vAR SD == within run SD`
:::



```{r}
#| label: run variance model

library(glmmTMB)
library(DHARMa)
# mean_prop = mean across simulation replicates

hist(hyb_run_mean_100$mean_HI)

## select only one age class to have 1 line = 1 run
# stage 1
m1 <- glmmTMB(mean_HI ~ selection * configuration * ori_proportion + (1 | run),family=gaussian, data = subset(hyb_run_mean_100, stage == 1))
summary(m1)
simulateResiduals(m1) |> plot()

# stage 2
m2 <- glmmTMB(mean_HI ~ selection * configuration * ori_proportion + (1 | run),family=gaussian, data = subset(hyb_run_mean_100, stage == 2))
summary(m2)
simulateResiduals(m2) |> plot()

# stage 3 ----> not good residuals are not normal ??
m3 <- glmmTMB(mean_HI ~ selection * configuration * ori_proportion + (1 | run),family=gaussian, data = subset(hyb_run_mean_100, stage == 3))
summary(m3)
simulateResiduals(m3) |> plot()


## repeat for 1000 ?

```

From the variance of the random effect (run) it looks like the variation among biological replicates is small compared to the remaining variation across parameter combinations. 

= Initial planting stochasticity has negligible effect on mean HI at generation 100 (stage 1). The system is robust to variation in initial spatial configuration (at least by generation 100). 

Other observations: 

-   proportion of orientalis has the stronger effect on the results (as expected)

-   effect of configuration: one cluster has higher HI, multicluster has moderately higher HI, transect not very different

-   the effect of selection doesn't seem to be significant at generation 100 compared to spatial configuration and initial orientalis proportion == **spatial structure matters more than selection???**


### Spatial distribution of hybrid proportion

Here we calculate the proportion of overall hybrids (everything that is not pure) per patch and plot across the landscape.

NB: this is not distinguishing the different hybrid classes.

```{r}
#| label: overall hybrid proportions per patch
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"


library(dplyr)
library(ggplot2)
library(terra)

## proportion of overall hybrids per patch 

hyb_prop_patch <- dt[,.( N = .N,
     N_hybrid = sum(P1 > -1 & P1 < 1),
     prop_hybrids = mean(P1 > -1 & P1 < 1)), 
  by = .(configuration, ori_proportion,selection,generation, replicate, stage, k, b, pop) ]

# importing grid 
grid <- vect("C:/Users/stefanin/Dropbox/WSL_PhD/Projects/Hybridization2/Input_files/Grid_4x4m_100x100m.shp")
grid_r <- terra::rast(ext = ext(grid), resolution = 4, crs = "EPSG:3035")
# grid dimensions 
n_rows = dim(grid_r)[1]
n_cols = dim(grid_r)[2]

## check the proportion of hybrids (for one stage and one replicate)
ggplot(subset(hyb_prop_patch, stage ==1 & replicate ==1),
         aes(x = (pop - 1) %% n_rows + 1, 
             y = n_cols - ((pop - 1) %/% n_cols + 1), 
             fill = as.numeric(prop_hybrids))) +
    geom_raster() +
    scale_fill_viridis_c(option = "viridis", direction = -1) +
    facet_grid(configuration + ori_proportion ~ generation) + 
    labs(x = "", y = "", fill = "Proportion of hybrids") +
    theme_void() +
    theme(
      aspect.ratio = 1,
      legend.position = "bottom",
      panel.spacing.x = unit(1, "lines"),
      panel.spacing.y = unit(0.1, "lines"),
      strip.text = element_text(size = 10))

```

### Hybrid classes

Below we group the hybrids into discrete classes (but very messy and not informative) - not very good plot + stacked bar chart after summarising across replicates (on average, what fraction of individuals are F1 / BC / etc?)

```{r}
#| eval: false
#| output: false
#| echo: true
#| code-fold: true
#| code-summary: "Expand for full code"


## hybrid classes
dt[, hybrid_class := fcase(
  P1 == 1,                "Pure orientalis",
  P1 == -1,                "Pure sylvatica",
  abs(P1) <= 0.2,              "F1", # between 0 and + - 0.2
  
  P1 <= 0.8 & P1 > 0.2,       "F2 or BC orientalis", 
  P1 >= -0.8 & P1 < -0.2,     "F2 or BC sylvatica", 
  default = "introgressed"
)]

## proportion per simulation 
hyb_classes <- dt[,
  .N,by = .(configuration, ori_proportion,selection,generation, replicate, stage, k, b, hybrid_class)
][
  , prop := N / sum(N),
  by = .(configuration,ori_proportion,selection,generation, replicate, stage, k, b)
]

## plot for each stage/generation the proportion of classes
# subset either generation or stage
ggplot(subset(hyb_classes, stage == 1), 
       aes(x = factor(generation), y = prop, group = interaction(hybrid_class, factor(generation)), color = factor(hybrid_class), fill = factor(hybrid_class)))+theme_bw()+
  geom_boxplot()+ 
  facet_grid(configuration+selection~ k+ori_proportion) 


### otherwise
# summarize across replicates -- not very good
hyb_class_summary <- hyb_classes[
  ,
  .(
    mean_prop = mean(prop),
    sd_prop   = sd(prop)
  ),
  by = .(configuration, ori_proportion, selection, generation, stage, k, b, hybrid_class)
]

ggplot(hyb_class_summary ,
       aes(x = factor(generation),
           y = mean_prop,
           fill = hybrid_class)) +
  geom_col() +
  facet_grid(configuration  ~ stage + ori_proportion) +
  labs(y = "Mean proportion", x = "Year") +
  theme_bw()
```
